---
title: "DETSI Moreton Bay Report"
author: "EP Consulting and Dumelle Consulting"
format:
  pdf:
    toc: true
    cite-method: natbib
    biblio-style: plainnat
  html:
    default
execute: 
  echo: false
  message: false
  warning: false
  error: true
  cache: true
embed-resources: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r}
# loading software
library(tidyverse)
library(here)
library(ranger)
library(sf)
library(knitr)
library(broom)
library(spmodel)
library(pdp)
library(emmeans)
library(janitor)
library(kableExtra)
```


# Background

Accurately characterizing drivers and resilience of seagrass cover is essential to improving the ecological health of the Moreton Bay and Wide Bay ecosystems. Over the past few decades, various scientific efforts have informed our current understanding of trends, behaviors, and flood responses of seagrass cover in both Moreton Bay and Wide Bay [@preen1995flood; @mcmahon2005herbicide; @york2022post] [@abal1996seagrass; @udy1997growth; @samper2016organic; @ovsyanikova2026spatial] as well as Australia more widely [@kilminster2015unravelling; @gilby2018seagrass; @o2018seagrass]. Remote sensing has offered new capabilities for seagrass cover mapping [@roelfsema2009integrated], while innovative statistical approaches have offered novel methods for studying resilience at site-specific scales [@gilby2023drivers]. Our work contributes to this body of literature by applying spatially explicit statistical models [@cressie1993statistics] to study seagrass cover drivers, resilience, and recovery in both Moreton Bay and Wide Bay. Specifically, we provide and contextualize answers to four specific research questions, applicable to both Moreton Bay and Wide Bay:

1. Can we identify water quality covariates that act as triggers for seagrass cover management responses? Are these triggers different based on whether they result from a specific event (e.g., flood) or are ambient?
2. Did seagrass cover improve after the flood event(s)? If so, how and where?
3. Can we identify site-specific resilience indicators, which describe where recovery is likely to be enhanced or impeded?
4. What factors contribute to seagrass resilience?

The spatially explicit statistical models (i.e., spatial models) we used to answer these research questions are related to commonly used linear and generalized linear models but incorporate _spatial dependence_ Spatial dependence measures how similar two sites are based on their distance. Formally, we can build spatial dependence directly into spatial models by leveraging Tobler's First Law of Geography [@tobler1970computer], which states succinctly that "nearby sites tend to be more similar than distant sites". Not only is this intuitive based on "first principles" perspective, but spatial models to provide many advantages for modeling spatial data compared to models that assume independence among observations (i.e., nonspatial models). First, nonspatial models tend to underestimate uncertainty, yielding covariate standard errors and p-values for covariates that are too small and corresponding confidence intervals that are too narrow. Second, spatial models improve prediction at unsampled locations by borrowing strength from nearby observations, yielding more precise predictions and corresponding standard errors that vary by location. Third, spatial models decompose model error into separate spatial and nonspatial components, informing residual and resilience structure in ways that nonspatial models cannot. Fourth, spatial models can provide ecologically meaningful insight into the spatial dependence structure of the process, informing relationships among measured and unmeasured variables as well as future sampling plans. @zimmerman2024spatial provide a thorough review of spatial models for ecological and environmental data, providing more detail and justification for the claims made above.

Spatial models offer many benefits over commonly used machine learning algorithms like random forests [@breiman2001random; @cutler2007random; @james2017introduction; @kuhn2022tidy]. First, traditional machine learning algorithms do not incorporate spatial dependence and, by consequence, often perform worse than spatial models at characterizing important covariates and predicting at unobserved locations [@fox2020comparing]. Second, machine learning algorithms struggle to quantify uncertainty in a principled manner, something spatial models excel at. Third, covariate inference is more straightforward using spatial models, as they naturally provide p-values that enable structured testing of hypotheses. Fourth, spatial models provide a rich set of diagnostics like leverage and Cook's distances [@montgomery2021introduction], quantities not generally defined for machine learning algorithms. This is not to say that machine learning algorithms cannot be very useful in ecological data analyses, but rather than spatial models are often a more effective tool for modeling spatial data. 

Though there are many benefits of spatial models over machine learning algorithms, the reverse is also true. Machine learning algorithms much more simply capture complex nonlinearities "out-of-the-box", bypassing the tuning often required for spatial models that involve nonlinear components like splines. This can be especially useful for ecological data, which are often full of nonlinear relationships [@fox2017assessing]. Machine learning algorithms also typically rely on few, if any, distributional assumptions and can be more robust against unusual observations like outliers [@hastie2009elements]. Machine learning algorithms also tend to be much mroe computational efficient than spatial models. Importantly, machine learning algorithms and spatial models can be complementary tools that help shed insight into complicated ecological processes (something we reiterate in the upcoming analyses). An exciting and developing area of statistical research involves incorporating spatial modeling strategies directly into machine learning algorithms [@saha2023random; @heaton2025scalable].

The first spatially explicit model we define is the spatial linear model. The spatial linear model is a _mixed-effects model_ [@pinheiro2000mixed; @bolker2015linear] written as

$$
y_i = \beta_0 + \beta_1x_{1, i} + \beta_2x_{2, i} + ... + \beta_p x_{p, i} + \tau_i + \epsilon_i,
$$ {#eq-splm1}
where $i = 1, 2, ..., n$ indexes the $n$ observations. The quantity $y_i$ is the response variable (for observation $i$), the $\beta$ parameters are slope parameters controlling the covariates, $x_{j, i}$ (for $j = 1, 2, ..., p$). The random error $\tau_i$ is a spatial random error, and the random error $\epsilon_i$ is the nonspatial (i.e., independent) random error. The spatial random error is spatially structured, capturing spatial residual effects like environmental gradients we do not have data to represent. The nonspatial random error is not spatially structure and captures nonspatial residual effects like measurement error and microscale variation. The difference between a traditional (i.e., nonspatial) linear model and a spatial linear model is the inclusion of $\tau_i$ (in the spatial linear model). The spatial random error implied by $\tau_i$ is governed by a _spatial covariance function_. The spatial covariance function allows the model to incorporate spatial dependence, yielding the improved model performance previously described. @fig-spcov_fns shows three different spatial covariance functions that describe different ways the spatial covariance decays with distance. Model parameters and standard errors are estimated from data using likelihood-based [@patterson1971recovery; @harville1977maximum; @wolfinger1994computing] or semivariogram-based [@cressie1985fitting; @curriero1999composite] estimation methods. Often, the spatial linear model in @eq-splm1 in is written more compactly using _matrix notation_:
$$
\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
$$ {#eq-splm2}
where $\mathbf{X}$ is a matrix that holds the $j$ covariates for each of $n$ observations, and $\mathbf{y}$, $\boldsymbol{\beta}$, $\boldsymbol{\tau}$, and $\boldsymbol{\tau}$ are vectors that contain each of the $n$ or $j$ quantities from @eq-splm1. The spatial linear model can be adjusted to account for nonspatial random effects just as one would add random effects to a nonspatial linear mixed model:
$$
\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \mathbf{Z}\mathbf{u} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
$$ {#eq-splm3}
where $\mathbf{Z}$ is a random effect design matrix (dimension $n \times k$) that indexes each of the $k$ random effects in $\mathbf{u}$. For more information about specific forms of various spatial covariance functions, see @schabenberger2017statistical and @zimmerman2024spatial.

```{r}
#| label: fig-spcov_fns
#| fig-cap: "Three spatial covariance functions that represent different spatial dependence behavior."
range <- 50
spcor_dat <- tibble(
  h = seq(0, range*2),
  Exponential = exp(-h/range),
  Gaussian = exp(-h^2/range^2),
  Spherical = (1 - 1.5*h/range + 0.5*h^3/range^3)*(h <= range)
) %>%
  rename(Distance = h) %>%
  pivot_longer(cols = c(-Distance), names_to = "Function", values_to = "Spatial Covariance")
ggplot(spcor_dat, aes(x = Distance, y = `Spatial Covariance`, color = Function, linetype = Function)) +
  geom_line(linewidth = 2) +
  theme_bw(base_size = 14)
```

Like the traditional linear model, the spatial linear model can capture complex nonlinearities in the data through the use of interaction effects [@faraway2002practical; @lenth2025emmeans], splines [@chambers1992statistical], additive models [@wood2017generalized; @pedersen2019hierarchical], and penalized regression models like the ridge and lasso [@tibshirani1996regression]. Generally, this is accomplished by leveraging _basis functions_ that break down nonlinear behavior into a set of additive linear components that may include penalty terms which guard against overfitting. Furthermore, there is no assumption that $y_i$ itself is normally distributed; rather, distributional assumptions are placed on the random errors after incorporating the slope parameters. This clarification implies the spatial (and nonspatial) linear models can be quite effective at modeling a range of data that are not normally distributed. For highly nonnormal data, tools like generalized linear models, which we discuss later, are helpful. Finally, under general conditions, the slope parameters of the spatial (and nonspatial) linear models are approximately normally distributed with a large sample size (see @casella2024statistical for a general justification via the Central Limit Theorem and @zhang2005towards for nuances regarding spatial data). This is beneficial because it justifies valid hypothesis testing on the slope coefficients (often of primary ecological interest), no matter the distribution of the data (given certain assumptions are satisfied).

One of the most useful applications of the spatial linear model is predicting the response variable at unobserved locations. A very useful spatial prediction method called Kriging has a rich theoretical justification [@cressie1990origins; @stein1999interpolation]. Kriging is the spatial analogue to best linear unbiased prediction (BLUP) for mixed models [@henderson1975best]. Via Kriging, the spatial linear model produces BLUPs at each unobserved location of interest, alongside prediction intervals to characterize uncertainty. While useful, sometimes the goal is to create a BLUP for an entire region of interest. This technique, which builds upon the theoretical foundations of Kriging, is called block Kriging [@ver2002sampling; @cressie2006block]. The intuition is to make point predictions at a very dense grid in the study region and then average these predictions, appropriately characterizing uncertainty. Together, Kriging and block Kriging provide a complete set of tools for spatial prediction, given the fit of a spatial linear model.

Spatial generalized linear models are an extension of spatial linear models to highly nonnormal data [@christensen2002bayesian; @zhang2002estimation; @diggle2007model; @bonat2016practical]. @verhoef2024marginal proposed a novel application of the Laplace approximation to enable restricted maximum likelihood estimation of spatial generalized linear models. Their model formulation builds upon @eq-splm1 and is given by
$$
g(\mu_i) = \beta_0 + \beta_1x_{1, i} + \beta_2x_{2, i} + ... + \beta_p x_{p, i} + \tau_i + \epsilon_i,
$$ {#eq-spglm1}
where $g(\mu_i)$ is a "link function" that links the distribution of $y_i$ to a linear function of its mean, $\mu_i$. Common link functions include the log link for count and skewed data and the logit link for binomial and proportion data (@tbl-glm-families). Using the same argument as for spatial linear models, nonspatial random effects can be added to spatial generalized linear models.

```{r}
#| label: tbl-glm-families
#| tbl-cap: "Common generalized linear model families, link functions, and appropriate data types."

glm_table <- data.frame(
  Family = c("Binomial", "Poisson", "Negative Binomial",
             "Beta", "Gamma", "Inverse Gaussian"),
  Link_Function = c("$f(\\mu) = \\log(\\mu / (1 - \\mu))$",
                    "$f(\\mu) = \\log(\\mu)$",
                    "$f(\\mu) = \\log(\\mu)$",
                    "$f(\\mu) = \\log(\\mu / (1 - \\mu))$",
                    "$f(\\mu) = \\log(\\mu)$",
                    "$f(\\mu) = \\log(\\mu)$"),
  Link_Name = c("Logit", "Log", "Log", "Logit", "Log", "Log"),
  Data_Type = c("Binary; Binary Count", "Count", "Count",
                "Proportion", "Skewed", "Skewed")
)

kable(glm_table,
      col.names  = c("Family", "Link Function", "Link Name", "Data Type"),
      escape     = FALSE,
      booktabs   = TRUE)
```

Spatial models are complicated, both theoretically and computationally. This has contributed to their relative lack of use in the ecological community. However, recent advances in open-source software have made spatial models more accessible to ecologists who have strong subject matter expertise but may lack sufficient statistical background or computational skills to implement them from scratch. One such advancement is the `spmodel` **R** package [@dumelle2023spmodel], which emulates base-**R** functions like `lm()` and `glm()` to account for spatial dependence via `splm()` and `spglm()`, respectively. Familiar **R** functions like `summary()` and `predict()` acn be used directly on models fit via `spmodel`. Importantly, `spmodel` also has tools for spatial random forest residual modeling [@fox2020comparing], applications to large data of several thousand observations via spatial indexing [@ver2023indexing], and cross validation [@stone1974cross; @roberts2017cross].

# Moreton Bay

## The Data

```{r}
#| label: read-mb
dat <- read_sf(here("data", "MBsub_03-31-2026.gpkg"))

# read in bay shapefiles
bay <- read_sf(here("data", "Moreton_Bay_marine_park_zoning_2008.shp")) %>%
  st_transform(crs = 3113)
```

We started with just over 10,000 seagrass cover observations in Moreton Bay across three financial years (2020-2021, 2022-2023, and 2023-2024). eReefs data [@steven2019ereefs] with modeled water quality covariates related to turbidity and salinity were only available at 7,657 observaations. Because one of our research questions involved connecting water quality covariates to management actions, we proceed with this subset of 7,657 observations for modeling. Of the 7,657 observations, 1,995 were in 2020-2021 (pre-flood), 3,329 were in 2022-2023, and 2,333 were in 2023-2024 (@tbl-n_fy_mb). 

```{r}
#| label: tbl-n_fy_mb
#| fig-cap: "Total seagrass observations in Moreton Bay by financial year."
dat %>%
  tabyl(fy) %>%
  adorn_totals(where = "row") %>%
  adorn_pct_formatting() %>%
  kable(
    col.names = c("Financial Year", "Sample Size", "Percent (of Total)"),
    format.args = list(big.mark = ",")
  )
```

## Water Quality and Geographic Modeling

Total cover is a percentage that ranges from 0 (no seagrass) to 100 (seagrass entirely contained in an area). @fig-cover_fy_mb shows total cover by financial year throughout the Bay. Total cover tends to be largest along the eastern shoreline of the Bay and lowest in the middle of the Bay. @fig-cover_fy_mb also reveals that 2022-2023 had by far the most sampling in the middle of the Bay.


```{r}
#| label: fig-cover_fy_mb
#| fig-cap: "Total seagrass cover in Moreton Bay by financial year."
dat <- dat %>%
  arrange(total.cover, fy)
ggplot(dat) +
  geom_sf(data = bay, fill = "white") +
  geom_sf(data = dat, aes(color = total.cover), size = 0.8) +
  scale_color_viridis_c(name = "Cover") +
  scale_x_continuous(breaks = c(153.1, 153.3, 153.5)) +
  facet_wrap(~ fy) +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

We considered for modeling several water quality and geographic covariates. @tbl-wq_vars_mb summarize the water quality variables, which included both ambient and flood-related metrics related to turbidity, salinity, and temperature. The "Variable" column indicates the water quality variable turbidity, salinity, or temperature. The "Flood" variable indicates whether the measurement corresponds to a flood variable or not. The "Period" variable indicates the prior period (in days) relative to sampling or indicates whether the variable corresponds to a flood event. The "Type" column is the type of statistic, either a mean, minimum, maximum, number of days above (or below) a cutoff during "Period", or a maximum spell (i.e., duration) above (or below) a cutoff during "Period". The "Included" variable indicates whether the variable was included as a covariate in the final spatial model. We included only a few variables in the final model: mean turbidity in the 90 days prior to sampling, mean turbidity during the flood event, and mean temperature in the 90 days prior to sampling. We only included a single turbidity metric prior to sampling because these variables were all quite correlated with one another, making it hard for the model to tease apart signals. We believed 90 day mean turbidity prior to sampling to be the most accurate "running average" of turbidity prior to sampling (@fig-cover_turb_mb). We used a similar rationale to select only a single turbidity metric from the flood and a single temperature metric prior to sampling. We omitted a flood salinity metric primarily because we lacked data for salinity prior to sampling, thus it would be challenging to establish a "baseline" for salinity in order to tease apart prior to sampling effects from flood effects. We provide more detail regarding the modeling in @sec-mb_model.

```{r}
#| label: tbl-wq_vars_mb
#| tbl-cap: "Moreton Bay water quality variables considered for modeling."
turb_vars <- tibble(
  Variable = rep("Turbidity", 20),
  Flood = c(rep("No", 12), rep("Yes", 8)),
  Period = c(rep(c("7", "30", "90"), times = 4), rep("Flood", 8)),
  Type = c(rep("Mean", 3), rep("Above", 9), rep("Above", 3), rep("Duration", 3), c("Mean", "Max")),
  Cutoff = c(rep("NA", 3), rep(c("3", "5", "6"), each = 3), rep(c("3", "5", "6"), times = 2), rep("NA", 2))
) %>%
  mutate(Included = if_else(row_number() %in% c(3, 19), "Yes", "No"))
sal_vars <- tibble(
  Variable = rep("Salinity", 17),
  Flood = c(rep("No", 9), rep("Yes", 8)),
  Period = c(rep(c("7", "30", "90"), times = 3), rep("Flood", 8)),
  Type = c(rep("Below", 9), rep("Below", 3), rep("Duration", 3), c("Mean", "Min")),
  Cutoff = c(rep(c("20", "25", "30"), each = 3), rep(c("20", "25", "30"), times = 2), rep("NA", 2))
) %>%
  mutate(Included = if_else(row_number() %in% c(0), "Yes", "No"))
temp_vars <- tibble(
  Variable = rep("Temperature", 3),
  Flood = c(rep("No", 3)),
  Period = c(rep(c("7", "30", "90"), times = 1)),
  Type = c(rep("Mean", 3)),
  Cutoff = c(rep("NA", 3))
) %>%
  mutate(Included = if_else(row_number() %in% c(3), "Yes", "No"))
wq_vars <- bind_rows(turb_vars, sal_vars, temp_vars)
wq_vars %>%
  kable()
```

```{r}
#| label: fig-cover_turb_mb
#| fig-cap: "Total seagrass cover in Moreton Bay by turbidity class and financial year. Boxplot width is proportional to the number of observations in each x-axis level (within financial year). The y-axis is truncated at 50, though a few observations have higher total seagrass cover."
dat_turb <- dat %>%
  mutate(exceed_0.5 = factor(if_else(turb.90d.mn > 0.5, "Yes", "No"), levels = c("Yes", "No")))
ggplot(dat_turb, aes(x = exceed_0.5, y = total.cover)) +
  geom_boxplot(varwidth = TRUE) +
  facet_wrap(~ fy) + 
  labs(x = "Exceeds 0.5 NTU", y = "Total Cover") +
  coord_cartesian(ylim = c(0, 50)) +
  theme_bw(base_size = 14)
```


@tbl-geo_vars_mb summarizes the geographic variables, which have the following structure:

* The "Sediment Type" variable has six levels: Sand, Sandy Mud, Muddy Sand, Mud, Rock, and Other.
* The "Tidal Range" variable has three levels: Shallow Subtidal, Intertidal, and Deep Subtidal (@fig-cover_tidal_mb).
* The "Years Since Flood" is numeric from zero (pre flood) to 1.93 years post flood (February, 2024). 
* The "Nearest River" variable has six levels: Brisbane River, Caboolture River, Logan River, Ningi Creek, Pimpama Coomera and Nerang Rivers, and Pine River.
* The "Distance to (Nearest River) Outlet (in km)" variable is numeric from 0.24 to 34.97.
* The "Growing Season" variable has two levels: Growing (September - December) and Not Growing (January - August). 
* The EHMP subzone variable has eleven levels: Bramble Bay, Broadwater, Central Bay, Deception Bay, Eastern Banks, Moreton Bay Open Coastal Waters, North-Eastern Moreton Bay, Northern Banks Open Coastal Waters, South-Eastern Moreton Bay, Southern Bay, and Waterloo Bay (@fig-ehmp_subzone). 

```{r}
#| label: tbl-geo_vars_mb
#| fig-cap: "Moreton Bay geographic variables considered for modeling."
geo_vars <- tibble(
  Variable = c("Sediment Type", "Tidal Range", "Years Since Flood",
  "Nearest River", "Distance to Outlet", "Growing Season", "EHMP Subzone"), 
  Effect = c(rep("Fixed", 6), "Random"),
  Included = c(rep("Yes", 3), rep("No", 3), "Yes")
)
geo_vars %>%
  kable()
```

```{r}
#| label: fig-cover_tidal_mb
#| fig-cap: "Total seagrass cover in Moreton Bay by tidal class and financial year. Boxplot width is proportional to the number of observations in each x-axis level (within financial year)."
dat_tidal <- dat %>%
  mutate(depth.range = case_when(
    depth.range == "deep subtidal" ~ "DSt",
    depth.range == "shallow subtidal" ~ "SSt",
    depth.range == "intertidal" ~ "Int",    
  )) %>%
  mutate(depth.range = factor(depth.range, levels = c("Int", "SSt", "DSt")))
ggplot(dat_tidal, aes(x = depth.range, y = total.cover)) +
  geom_boxplot(varwidth = TRUE, size = 0.5) +
  facet_wrap(~ fy) + 
  labs(x = "Tidal Class", y = "Total Cover") +
  coord_cartesian(ylim = c(0, 20)) +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```


```{r}
#| label: fig-ehmp_subzone
#| fig-cap: "EHMP Subzones in Moreton Bay."
ggplot(dat) +
  geom_sf(data = bay, fill = "white") +
  geom_sf(data = dat, aes(color = ehmp.subzone), size = 0.5) +
  scale_color_discrete(name = "EHMP Subzone") +
  scale_x_continuous(breaks = c(153.1, 153.3, 153.5)) +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

From @tbl-geo_vars_mb, we kept the sediment type, tidal range, and years since flood variables as covariates in the model. The nearest river and distance to nearest outlet variables were removed, as they did not seem related to total cover. We removed growing season from the model after determining that it likely it not ecologically useful, as a simple month proxy likely does not adequately account for the complexities of the more nuanced aspects of seagrass growth with time. We used EHMP subzone as a random effect to account for within-subzone correlation that may impact the relationships among total cover and the remaining variables. We provide more detail regarding the modeling in @sec-mb_model.

### Model Fit {#sec-mb_model}

```{r}
#| label: tbl-mb_tc_splm_vars
#| fig-cap: "Variables in the final Moreton Bay total cover model, their type (numeric or categorical), and if categorical, the number of unique levels."
mb_tc_splm_vars <- tibble(
  Variable = c("90 Day Temperature", "90 Day Turbidity", "Mean Flood Turbidity/Years Since Flood^2", "Sediment Type", "Tidal Range"), 
  Type = c(rep("Numeric", 3), rep("Categorical", 2)),
  `No. Levels` = c(rep("NA", 3), "6", "3")
)
mb_tc_splm_vars %>%
  kable()
```

```{r}
#| label: fig-time_since_flood
#| fig-cap: "Time since the flood against the squared reciprocal of time since the flood."
dat_yr.flood <- dat %>%
  filter(yr.flood > 0) %>%
  mutate(yr.flood = seq(min(.$yr.flood), max(.$yr.flood), length.out = NROW(.))) %>%
  mutate(yr.flood.recip2 = 1/yr.flood^2)
ggplot(dat_yr.flood, aes(x = yr.flood, y = yr.flood.recip2)) +
  geom_line() +
  labs(x = "Years Since Flood", y = expression("1/(Years Since Flood)"^2)) +
  theme_bw(base_size = 12)
```

```{r}
# moreton bay tc model
mb_tc_splm_mod <- readRDS(here("models", "mb_tc_splm_mod.rds"))
```

To describe total seagrass cover in Moreton Bay, our final model was a spatial linear model (@eq-splm1) with the following explanatory variables: tidal range, sediment type, 90 day mean temperature (prior to sampling), pre- vs post-flood, years since the flood, 90 day turbidity (prior to sampling), and mean turbidity during the flood period scaled by time since the flood squared (@tbl-mb_tc_splm_vars). We scaled mean turbidity during the flood period so that the effect of turbidity during the flood event was allowed to decay with time since the flood (@fig-time_since_flood). Intuitively, this means that with time since the flood, the impact of turbidity during the flood event is lessened. Additionally, we used an exponential spatial covariance function and included a random effect for EHMP subzone. We fit the model using restricted maximum likelihood via the `splm()` function in `spmodel`.

@tbl-mb-covariates shows the results from an analysis of variance applied to the fitted model. All explanatory variables had large test statistics and small p-values ($p$-value $<$ 0.01); here are some relevant takeaways:

* Tidal Range: The estimated (marginal) average total cover is largest in the intertidal (26.89%), followed by shallow subtidal (16.51%) and deep subtidal (4.88%).
* Sediment Type: The estimated (marginal) average total cover is largest in the other (25.6%), followed by muddy sand (19.1%), sandy mud (17.8%), mud (14.3%), sand (14.2%), and rock (5.62%).
* 90 Day Temperature: A one-degree (C) increase in 90 day temperature is associated with a decrease in average total cover by 0.88%.
* Mean Flood Turbidity / (Years Since Flood)$^2$: An one-unit increase in mean turbidity during the flood is associated with a decrease in average total cover by 2.05%/(Years Since Flood)$^2$.
* 90 Day Turbidity: A one-unit increase in 90 day turbidity is associated with a decrease in average total cover by 7.18%.

The estimated marginal means for tidal range and sediment type were calculated using the `emmeans` **R** package [@lenth2025emmeans].

```{r}
#| label: tbl-mb-covariates
#| tbl-cap: "ANOVA table of covariates in the Moreton Bay water quality and geographic model."
mb_tc_splm_mod %>%
  anova() %>%
  tidy() %>%
  filter(effects != "(Intercept)") %>%
  arrange(desc(statistic)) %>%
  mutate(effects = c("Tidal Range", "Sediment Type", "90 Day Temperature", "Mean Flood Turbidity/Years Since Flood^2", "90 Day Turbidity")) %>%
  kable(digits = 3)
```


We may inspect model diagnostics to better understand the overall model fit. Here we discuss how to partition the model into different components of variability, infer decay behavior of spatial autocorrlation, analyze standardized residuals to assess model assumptions, and identify influential observations that have a sizable impact on model fit. From @tbl-mb_varcomp, we see that of the total model variability, 11.2% is attributable to the explanatory variables (this quantity is sometimes called the Pseudo R-squared), 31.6% is attributable to the spatial random error, 7.3% is attributable to the EHMP subzone, and the remaining variability is independent random error. The estimated exponential spatial range ($\phi$) is 745.5 meters, which suggests two total cover observations are approximately (spatially) uncorrelated at a distance of $3\phi = \approx 2$km, or two kilometers. While the raw residuals will inform our resilience modeling in @sec-mb_resilience, we can check certain model assumptions using the standardized residuals, which have been decorrelated. Generally, the standardized residuals are centered around zero, though there is a slight right skew suggesting some observed total cover values are larger than expected (according to the model). Observations are called "influential" if they highly influence model fit. Influence is often measured via Cook's distance, and no observations exceed the often-used Cook's distance threshold value of one [@cook1982residuals].

```{r}
#| label: tbl-mb_varcomp
#| tbl-cap: "Variance components in the Moreton Bay water quality and geographic model."
mb_tc_splm_mod %>%
  varcomp() %>%
  mutate(varcomp = c("Covariates", "Spatial Effect", "Independent Error", "EHMP Subzone")) %>%
  adorn_pct_formatting() %>%
  mutate(val = c(NA, coef(mb_tc_splm_mod, type = "spcov")[1:3])) %>%
  kable(digits = 1, col.names = c("Variance Component", "Percent (of Total Variability)", "Parameter Value"))
```

### Block Kriging

A tremendous benefit of using the spatial linear model is that block Kriging can be directly applied to predict the mean total seagrass cover bay-wide for any combination of explanatory variables and geographic regions, even if these variables were not directly used in modeling. @fig-bk_mb_ehmp and @fig-bk_mb_zone show that there are several EHMP subzones and conseravtion zones that have recovered to pre-flood levels (e.g., Deception Bay, Marine National Park Zone), while others have not (e.g., Southern Bay, General Use Zone). While EHMP subzone was included in the model as a random effect, management zone type was not -- via block Kriging, we can still quantify broad trends in total seagrass cover. @fig-bk_mb_tidal shows that total seagrass cover is recovering less effectively in the intertidal range than the shallow or deep subtidal ranges, while @fig-bk_mb_all shows that total seagrass cover is recovering bay-wide since the flood. Another useful application of block Kriging is scenario modeling, which can be used to characterize the impact of various future changes to broad-scale climatic patterns.

```{r}
#| fig-width: 8
#| fig-height: 10
#| label: fig-bk_mb_ehmp
#| fig-cap: "Average total cover predictions by EHMP subzone. Red dots indicate the mean from the raw data. Black dots indicate the predicted mean in the region alongside a prediction interval."
bk_mb_ehmp <- readRDS(here("models", "bk_mb_ehmp.rds"))
ggplot(bk_mb_ehmp, aes(x = fy, y = fit)) +
  geom_point() +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.3) +
  geom_point(aes(x = fy, y = raw_mean), color = "red") +
  facet_wrap(~ ehmp.subzone, ncol = 2) +
  theme_bw(base_size = 12)
```

```{r}
#| fig-width: 8
#| fig-height: 10
#| label: fig-bk_mb_zone
#| fig-cap: "Average total cover predictions by EHMP conservation zone. Red dots indicate the mean from the raw data. Black dots indicate the predicted mean in the region alongside a prediction interval."
bk_mb_zone <- readRDS(here("models", "bk_mb_zone.rds"))
ggplot(bk_mb_zone, aes(x = fy, y = fit)) +
  geom_point() +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.3) +
  geom_point(aes(x = fy, y = raw_mean), color = "red") +
  facet_wrap(~ zone.type, ncol = 2) +
  theme_bw(base_size = 12)
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| label: fig-bk_mb_tidal
#| fig-cap: "Average total cover predictions by EHMP tidal range. Red dots indicate the mean from the raw data. Black dots indicate the predicted mean in the region alongside a prediction interval."
bk_mb_tidal <- readRDS(here("models", "bk_mb_tidal.rds"))
ggplot(bk_mb_tidal, aes(x = fy, y = fit)) +
  geom_point() +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.3) +
  geom_point(aes(x = fy, y = raw_mean), color = "red") +
  facet_wrap(~ depth.range, ncol = 2) +
  theme_bw(base_size = 12)
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| label: fig-bk_mb_all
#| fig-cap: "Average total cover predictions by Moreton Bay. Red dots indicate the mean from the raw data. Black dots indicate the predicted mean in the region alongside a prediction interval."
bk_mb_all <- readRDS(here("models", "bk_mb_all.rds"))
ggplot(bk_mb_all, aes(x = fy, y = fit)) +
  geom_point() +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.3) +
  geom_point(aes(x = fy, y = raw_mean), color = "red") +
  theme_bw(base_size = 12)
```

### Alternative Approaches

In addition to the spatial linear model, we considered several different modeling approaches (all using the same explanatory variables): a nonspatial linear model, a spatial beta regression model, a spatial gamma regression model, and a random forest model. The nonspatial linear model provided a useful baseline to compare the spatial linear model against, elucidating the impact of spatial dependence on model fit. The spatial beta regression model is a generalized linear model used to characterize proportion data between zero and one. To adhere to this restriction, total cover was 1) scaled to a proportion and 2) if exactly zero (one), set to 0.01 (0.99).  The spatial gamma regression model is a generalized linear model used to characterize skewed positive data. To adhere to this restriction, if total cover was exactly zero, it was set to 0.01. Finally, the random forest provided a useful comparison to a machine learning approach, quite different than the aforementioned statistical approaches.

```{r}
mb_tc_lm_loocv <- readRDS(here("models", "mb_tc_lm_loocv.rds"))
mb_tc_splm_loocv <- readRDS(here("models", "mb_tc_splm_loocv.rds"))
mb_tc_spglm_beta_loocv <- readRDS(here("models", "mb_tc_spglm_beta_loocv.rds")) * 100 # to undo scaling
mb_tc_spglm_gamma_loocv <- readRDS(here("models", "mb_tc_spglm_gamma_loocv.rds"))
mb_tc_lm_loocv <- readRDS(here("models", "mb_tc_lm_loocv.rds"))
mb_tc_rf_oob <- readRDS(here("models", "mb_tc_rf_oob.rds"))
mb_cv <- bind_rows(mb_tc_lm_loocv, mb_tc_splm_loocv, mb_tc_spglm_beta_loocv,
                   mb_tc_spglm_gamma_loocv, mb_tc_rf_oob)
```

Importantly, all five approaches found important relationships among the explanatory variables and total cover, in the directions expected. From @tbl-mb_cv, the spatial linear model and random forest had the lowest out-of-sample root-mean-squared-prediction error, followed by the generalized linear models and then by the nonspatial linear model. Random forest partial dependence plots [@greenwell2017pdp], which characterize the marginal behavior of a variable in a random forest model, suggested linear effects of the model variables to be reasonable. Given this context, we chose the spatial linear model as the final model.
```{r}
#| label: tbl-mb_cv
#| tbl-cap: "Moreton Bay out of sample prediction metrics for the various approaches."
mb_cv %>%
  mutate(approach = c("LM", "SPLM", "SPGLM-Beta", "SPGLM-Gamma", "RF")) %>%
  select(approach, bias, RMSPE) %>%
  arrange(RMSPE) %>%
  kable(digits = 2, col.names = c("Approach", "MBias", "RMSPE"))
```

### Limitations and Future Suggestions

While the spatial linear model was useful for characterizing broad patterns related to tidal range, sediment type, temperature, and turbidity while accounting for spatial dependence, there are some limitations. First, the eReefs water quality data are modeled, not directly observed in the field, potentially propogating modeling error. Second, there are many correlated water quality variables, making it hard (for any model) to isolate the effects of each component of water quality pre- and post-flood. Third, there is a lack of baseline mean salinity in the 90 days prior to sampling, which makes it challenging to directly study the effect of mean salinity during the flood period. Fourth, there is a lack of data immediately after the flood -- the first post-flood sampling effort occurred four months after the flood. Fifth, there is a lack of addition data sources that could help quantify the impact of various explanatory variables on total pre and post-flood, potentially varying across both space and time. Sixth, the locations at which data were collected were not randomly sampled. Randomly sampling sites helps to guard against sampling preferentially [@diggle2010preferential] and provides more robust and representative data sources [@dumelle2022comparison].

## Resilience Modeling {#sec-mb_resilience}

### Model Fit

### Alternative Approaches

## References

::: {#refs}
:::

# Appendix

