---
title: "Moreton Bay Seagrass Resilience Modeling"
subtitle: "DRFA 21/22 Biodiversity Conservation Marine Final Report Project"
author:
  - name: Michael Dumelle (Dumelle Consulting)
    affiliation: Dumelle Consulting
  - name: Erin E. Peterson (EP Consulting)
    affiliation: EP Consulting
date: "May 11, 2026"
format:
  pdf:
    toc: true
    cite-method: natbib
    biblio-style: plainnat
    number-sections: true
    keep-tex: true
    keep-md: true
  html:
    toc: true
    cite-method: natbib
    biblio-style: plainnat
    number-sections: true
    keep-tex: true
    keep-md: true
execute: 
  echo: false
  message: false
  warning: false
  error: true
  cache: false
embed-resources: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r}
# loading software
library(tidyverse)
library(here)
library(ranger)
library(sf)
library(knitr)
library(broom)
library(spmodel)
library(pdp)
library(emmeans)
library(janitor)
library(kableExtra)
library(splines)
library(patchwork)
library(xtable)
```

# Introduction

Over the past few decades, various scientific studies have informed our current understanding of trends, behaviors, and flood responses of seagrass cover in Moreton Bay [@abal1996seagrass; @udy1997growth; @samper2016organic; @ovsyanikova2026spatial]  as well as Australia more widely [@kilminster2015unravelling; @gilby2018seagrass; @o2018seagrass]. Remote sensing has offered new capabilities for seagrass cover mapping [@roelfsema2009integrated], while innovative statistical approaches have offered novel methods for studying resilience at site-specific scales [@gilby2023drivers]. Our work adds to this body of literature by applying spatially explicit statistical models [@cressie1993statistics] to study the drivers, resilience, and recovery of total percent seagrass cover in Moreton Bay and Wide Bay [@dumelle2026wide]. Specifically, we provide and contextualize answers to three specific research questions, applicable to Moreton Bay:

1. Can we identify water quality covariates that act as triggers for seagrass cover management responses? Are these triggers different based on whether they result from a specific event (e.g., flood) or ambient conditions?
2. Did seagrass cover improve after the flood event(s)? If so, how and where?
3. Can we identify site-specific resilience indicators, which describe where recovery is likely to be enhanced or impeded? If so, what factors contribute to seagrass resilience?


# Seagrass Cover Model
## Data

```{r}
#| label: read-mb
dat <- read_sf(here("data", "MBsub_03-31-2026.gpkg"))

# read in bay shapefiles
bay <- read_sf(here("data", "Moreton_Bay_marine_park_zoning_2008.shp")) %>%
  st_transform(crs = 3113)
```

The seagrass cover dataset was provided by Science Under Sail Australia (SUSA; https://www.scienceundersail.com.au/). It included 7657 observations across three financial years (2020-2021, 2022-2023, and 2023-2024), where water quality and geographic covariates were also available. Of the 7657 observations, 1995 were collected in 2020-2021 (pre-flood), 3329 in 2022-2023 (1-year post-flood), and 2333 in 2023-2024 (2-years post-flood) (@tbl-n_fy_mb). There were six seagrass species found in the dataset: *Halophila decipiens* (HD), *Halophila spinulosa* (HS), *Halophila ovalis* (HO), *Halodule uninervis* / *Zostera muelleri* (ZMHU), *Syringodium isoetifolium* (SI) and *Cymodocea serrulata* (CS).

```{r}
#| label: tbl-n_fy_mb
#| tbl-cap: "The sample size and percent of total sample size for total percent seagrass cover observations in Moreton Bay, by financial year."
#| results: asis
dat %>%
  st_drop_geometry() %>%
  tabyl(fy) %>%
  adorn_totals(where = "row") %>%
  adorn_pct_formatting() %>%
  rename(`Financial Year` = fy, `Sample Size` = n, `Percent of Total` = percent) %>%
  xtable() %>%
  print(include.rownames = FALSE, floating = TRUE, booktabs = TRUE, comment = FALSE)
```

Seagrass cover is a percentage ranging from 0 (no seagrass) to 100 (complete seagrass cover). Total cover tends to be highest near the eastern shoreline of the Bay and lowest in the middle (@fig-cover_fy_mb). Note that 2022-2023 (post-flood) had the the largest number of seagrass cover samples, as well as the only samples collected in the middle of the Bay.

```{r}
#| fig-width: 8
#| fig-height: 8
#| label: fig-cover_fy_mb
#| fig-cap: "Total percent seagrass cover in Moreton Bay by financial year."
dat1 <- dat %>%
  arrange(total.cover, fy)
ggplot(dat1) +
  geom_sf(data = bay, fill = "white") +
  geom_sf(data = dat1, aes(color = total.cover), size = 0.8) +
  scale_color_viridis_c(name = "Cover") +
  scale_x_continuous(breaks = c(153.1, 153.3, 153.5)) +
  facet_wrap(~ fy) +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

EcoFutures Consulting used eReefs data [@steven2019ereefs] to generate a broad suite of ambient and flood-related turbidity, salinity, and temperature variables at biologically and ecologically relevant temporal windows (@tbl-wq_vars_mb). Ambient water quality variables were based on the mean, minimum, and maximum values for three periods: 7, 30, and 90 days prior to sampling. Ambient variables were also calculated representing the frequency of days water quality thresholds were exceeded (above or below) and the maximum spell (i.e., duration) of days those thresholds were continuously exceeded during the time periods. Similar water quality variables representing flood-event conditions were also calculated for a 30-day flood period. 

```{r}
#| label: tbl-wq_vars_mb
#| tbl-cap: "Moreton Bay water quality variables considered in the exploratory analyses, including the variable name and measurement unit (in parentheses), the type of water-quality variable (Ambient, Flood), the time period in days, statistic calculated, and the threshold value used (if any)."
#| results: asis
turb_vars <- tibble(
  Variable = rep("Turbidity (NTU)", 20),
  Type = c(rep("Ambient", 12), rep("Flood", 8)),
  Period = c(rep(c("7", "30", "90"), times = 4), rep("30", 8)),
  Statistic = c(rep("Mean", 3), rep("Frequency", 9), rep("Frequency", 3), rep("Duration", 3), c("Mean", "Max")),
  Threshold = c(rep("", 3), rep(c("> 3", "> 5", "> 6"), each = 3), rep(c("> 3", "> 5", "> 6"), times = 2), 
             rep("", 2))
)

sal_vars <- tibble(
  Variable = rep("Salinity (PSU)", 17),
  Type = c(rep("Ambient", 9), rep("Flood", 8)),
  Period = c(rep(c("7", "30", "90"), times = 3), rep("30", 8)),
  Statistic = c(rep("Frequency", 9), rep("Frequency", 3), rep("Duration", 3), c("Mean", "Min")),
  Threshold = c(rep(c("< 20", "< 25", "< 30"), each = 3), rep(c("< 20", "< 25", "< 30"), times = 2), rep("", 2))
)

temp_vars <- tibble(
  Variable = rep("Temperature (deg C)", 3),
  Type = c(rep("Ambient", 3)),
  Period = c(rep(c("7", "30", "90"), times = 1)),
  Statistic = c(rep("Mean", 3)),
  Threshold = c(rep("", 3))
)

wq_vars <- bind_rows(turb_vars, sal_vars, temp_vars) %>%
  group_by(Variable, Type, Statistic, Threshold) %>%
  summarise(Period = paste(sort(unique(Period)), collapse = ", "),
            .groups = "drop") %>%
  mutate(
    Variable = factor(Variable, levels = c("Turbidity (NTU)", "Salinity (PSU)", "Temperature (deg C)")),
    Type = factor(Type, levels = c("Ambient", "Flood")),
    Statistic = factor(Statistic, levels = c("Min", "Mean", "Max", "Frequency", "Duration"))
  ) %>%
  arrange(Type, Variable, Statistic) %>%
  select(Variable, Type, Period, Statistic, Threshold)

wq_vars %>%
  xtable() %>%
  print(include.rownames = FALSE, floating = TRUE, booktabs = TRUE, comment = FALSE)
```

Additional covariates were provided by EcoFutures Consulting representing geographic features, proximity to rivers, and depth, as well as the years since the flood and seagrass growing season. These variables have the following structure:

* Sediment Type includes six categorical levels: Sand, Sandy Mud, Muddy Sand, Mud, Rock, and Other.
* Tidal Range has three categorical levels: Shallow Subtidal, Intertidal, and Deep Subtidal.
* Years Since Flood is numeric ranging from zero (pre flood) to 1.93 years post-flood (February, 2024). 
* Nearest River includes six categorical levels: Brisbane River, Caboolture River, Logan River, Ningi Creek, Pimpama Coomera and Nerang Rivers, and Pine River.
* Distance to (Nearest River) Outlet is numeric and ranges from 0.24 to 34.97 kilometers.
* Growing Season is a categorical variable with two levels: Growing (September - December) and Not Growing (January - August). 
* EHMP subzone has eleven categorical levels: Bramble Bay, Broadwater, Central Bay, Deception Bay, Eastern Banks, Moreton Bay Open Coastal Waters, North-Eastern Moreton Bay, Northern Banks Open Coastal Waters, South-Eastern Moreton Bay, Southern Bay, and Waterloo Bay (@fig-ehmp_subzone). 

```{r}
# #| label: tbl-geo_vars_mb
# #| tbl-cap: "Moreton Bay geographic variables considered for modeling."
# #| tbl-pos: 'h'
# geo_vars <- tibble(
#   Variable = c("Sediment Type", "Tidal Range", "Years Since Flood",
#   "Nearest River", "Distance to Outlet", "Growing Season", "EHMP Subzone"), 
#   Effect = c(rep("Fixed", 6), "Random"),
#   Included = c(rep("Yes", 3), rep("No", 3), "Yes")
# )
# geo_vars %>%
#   kable()
```


```{r}
#| fig-width: 8
#| fig-height: 6
#| label: fig-ehmp_subzone
#| fig-cap: "Observations and their EHMP Subzones in Moreton Bay."
ggplot(dat) +
  geom_sf(data = bay, fill = "white") +
  geom_sf(data = dat, aes(color = ehmp.subzone), size = 0.5) +
  scale_color_discrete(name = "EHMP Subzone") +
  scale_x_continuous(breaks = c(153.1, 153.3, 153.5)) +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```



## Exploratory Analyses

Extensive exploratory analyses were undertaken to 1) understand the data structure, 2) visualize spatial and temporal patterns in the data, 3) identify outliers/errors, 4) examine correlations between potential covariates and the seagrass cover, as well as multicollinearity among covariates, 5) compare the fit of various models, and 6) evaluate model assumptions.

Results revealed substantial collinearity among ambient and flood-related water-quality variables, as expected given their nested temporal windows and threshold values. We retained only one variable from each correlated group and prioritized sediment type, tidal range, turbidity, temperature, and time since flood when making these decisions. We also removed other variables showing no meaningful association with seagrass cover or where the relationship lacked biological plausibility. This included all of the salinity variables, which had little variability and did not exhibit strong relationships with seagrass cover. Geographic variables exhibited less collinearity than water quality variables; however, we removed the distance to the nearest river outlet and the nearest river variables due to weak associations with seagrass cover after accounting for other model covariates. Growing season was excluded because a simple monthly proxy inadequately captures seasonal seagrass growth dynamics.

The final set of potential covariates considered in the modelling included mean turbidity in the 90 days prior to sampling, mean turbidity during the flood event, mean temperature in the 90 days prior to sampling, sediment type, tidal range, years since flood, and EHMP subzone. 

## Statistical Modeling {#sec-mb_model}

We considered multiple modeling approaches in this study, which each accounted for the data distribution, nonlinear relationships between seagrass cover and covariates, and spatial dependence inherent in biological data collected through space and time to different degrees (Appendix 1). We used the full set of potential covariates to fit a nonspatial linear model, spatial linear model (i.e., spatial model), spatial beta regression model, spatial gamma regression model, and a random forest model, comparing the model fit of each.

There were several reasons for fitting these initial models. The nonspatial linear model provided a useful baseline to compare the spatial linear model against, elucidating the impact of spatial dependence on model fit. The spatial beta regression model is a generalized linear model used to characterize bounded proportional data ranging between zero and one. To adhere to this restriction, total seagrass cover was 1) scaled to a proportion and 2) if exactly zero or one, set to 0.01 or 0.99, respectively. The spatial gamma regression model is a generalized linear model used to characterize skewed positive data. To adhere to this restriction, a value of 0.01 was added to seagrass cover observations that were equal to zero. Finally, the random forest model provided a useful assessment of a machine learning approach, which requires fewer distributional assumptions than the aforementioned statistical approaches and more naturally describes nonlinear relationships between variables (Appendix 1). We also characterized random forest effects using partial dependence plots [@greenwell2017pdp], which isolate the marginal contribution of a variable while holding other variables at their observed values. Out-of-sample predictions were generated for the four statistical models using leave-one-out cross validation, while out-of-bag validation was used for the random forest model. These out-of-sample predictions were then used to calculate the mean bias (MBias) and root-mean-square-prediction error (RMSPE) for each model. 

```{r}
mb_tc_lm_loocv <- readRDS(here("models", "mb_tc_lm_loocv.rds"))
mb_tc_splm_loocv <- readRDS(here("models", "mb_tc_splm_loocv.rds"))
mb_tc_spglm_beta_loocv <- readRDS(here("models", "mb_tc_spglm_beta_loocv.rds")) * 100 # to undo scaling
mb_tc_spglm_gamma_loocv <- readRDS(here("models", "mb_tc_spglm_gamma_loocv.rds"))
mb_tc_lm_loocv <- readRDS(here("models", "mb_tc_lm_loocv.rds"))
mb_tc_rf_oob <- readRDS(here("models", "mb_tc_rf_oob.rds"))
mb_cv <- bind_rows(mb_tc_lm_loocv, mb_tc_splm_loocv, mb_tc_spglm_beta_loocv,
                   mb_tc_spglm_gamma_loocv, mb_tc_rf_oob)
```

The results of this initial model comparison showed that the spatial linear model had the lowest RMSPE, followed by the random forest, generalized linear models, and the nonspatial linear model (@tbl-mb_cv). Mean bias (MBias) tended to be small relative to root-mean-squared-prediction error (RMSPE). The nonspatial and spatial linear models also had much lower mean bias values than the random forest model, while the two generalized linear models produced unacceptably biased predictions. Nevertheless, all five approaches identified similar relationships between seagrass cover and the covariates, in the directions expected. Random forest partial dependence plots [@greenwell2017pdp] suggested that linear effects of the model variables were generally reasonable. We chose to use a spatial linear modeling framework for the Moreton Bay seagrass model based on these results and the ease of interpretability. 

```{r}
#| label: tbl-mb_cv
#| tbl-cap: "Moreton Bay out of sample prediction metrics for the five modeling approaches, including mean bias (MBias) and root mean square prediction error (RMSPE). "
#| results: asis
mb_cv %>%
  mutate(approach = c("Linear Model", "Spatial Linear Model", "Spatial GLM - Beta", "Spatial GLM - Gamma", "Random Forest")) %>%
  select(approach, bias, RMSPE) %>%
  arrange(RMSPE) %>%
  rename(Approach = approach, MBias = bias) %>%
  xtable(digits = 3) %>%
  print(include.rownames = FALSE, floating = TRUE, booktabs = TRUE, comment = FALSE)
```

```{r}
#| label: fig-rf_turb
#| fig-cap: "Random forest partial dependence for seagrass cover as a function of mean turbidity in the 90 days prior to sampling."
#| out-width: "50%"
mb_tc_rf_mod <- readRDS(file = here("models", "mb_tc_rf_mod.rds"))
pred.grid <- tibble(turb.90d.mn = seq(0, 1, length.out = 4))
pd <- partial(mb_tc_rf_mod, pred.var = "turb.90d.mn", pred.grid = pred.grid) 
p1 <- ggplot(pd, aes(x = turb.90d.mn, y = yhat)) +
  geom_line(linewidth = 2) +
  labs(x = "90 Day Mean Turbidity", y = "Seagrass Cover") +
  theme_bw(base_size = 14)
p1
```

A suite of spatial linear models (Appendix 1, @eq-splm1), with exponential spatial covariance functions and a random effect for EHMP subzone, were fit using the `splm()` function in `spmodel` @dumelle2023spmodel. Parameters were estimated using restricted maximum likelihood and models primarily compared using leave-one-out cross validation; though AIC or BIC could also be used if parameters were estimated using maximum likelihood. In addition, the relationships between seagrass cover and covariates needed to be plausible. 

The final total seagrass cover model for Moreton Bay included the following covariates: tidal range, sediment type, 90-day mean temperature (prior to sampling), 90 day turbidity (prior to sampling), and mean turbidity during the flood period scaled by time since the flood squared (@tbl-mb_tc_splm_vars). Mean turbidity during the flood period was scaled so that its effect decays with time since the flood. Intuitively, this means that as the time since the flood increases, the impact of turbidity during the flood event decreases. We also considered scaling by time since flood (not squared), which produced similar results.

```{r}
#| label: tbl-mb_tc_splm_vars
#| tbl-cap: "Covariates in the final Moreton Bay seagrass cover model, their type (numeric or categorical), and if categorical, the number of unique levels."
#| results: asis
mb_tc_splm_vars <- tibble(
  Variable = c("90 Day Temperature", "90 Day Turbidity", "Mean Flood Turbidity/Years Since Flood^2", "Sediment Type", "Tidal Range"), 
  Type = c(rep("Numeric", 3), rep("Categorical", 2)),
  `No. Levels` = c(rep("NA", 3), "6", "3")
)
mb_tc_splm_vars %>%
  xtable() %>%
  print(include.rownames = FALSE, floating = TRUE, booktabs = TRUE, comment = FALSE)

```

```{r}
# moreton bay tc model
mb_tc_splm_mod <- readRDS(here("models", "mb_tc_splm_mod.rds"))
```

```{r}
#| label: tbl-mb-fixed
#| tbl-cap: "Fixed effects table of covariates in the Moreton Bay water quality and geographic model."
#| results: asis
mb_tc_splm_mod %>%
  tidy() %>%
  mutate(term = c("Intercept (Mud/Deep Subtidal)", "Muddy Sand", "Other", "Rock", "Sand", "Sandy Mud",
                     "Intertidal", "Shallow Subtidal", "90 Day Temperature",
                     "90 Day Turbidity", "Mean Flood Turbidity/Years Since Flood^2")) %>%
  select(-std.error) %>%
  rename(Term = term, Estimate = estimate, z = statistic, p.value = p.value) %>%
  xtable(digits = 3) %>%
  print(include.rownames = FALSE, floating = TRUE, booktabs = TRUE, comment = FALSE)

```

@tbl-mb-fixed shows the fixed effects table from the fitted model. The reference group is the muddy and deep subtidal group, and the remaining sediment type and tidal range terms represent deviations from these baselines. For the numeric variables, all had small $p$-values ($p$-value $<$ 0.01). While we can assess the overall significance of the numeric variables @tbl-mb-fixed, we must perform an analysis of variance (ANOVA) to assess the overall significance of variables with many levels. @tbl-mb-covariates shows these ANOVA results for the sediment type and tidal range variables, which had small $p$-values ($p$-value $<$ 0.01). Together, these results indicate all variables had a significant relationship with total seagrass cover. Here are some relevant takeaways from the model fit:

* Tidal Range: The estimated (marginal) average total cover is largest in the intertidal (26.89%), followed by shallow subtidal (16.51%) and deep subtidal (4.88%).
* Sediment Type: The estimated (marginal) average total cover is largest in the other category (25.6%), followed by muddy sand (19.1%), sandy mud (17.8%), mud (14.3%), sand (14.2%), and rock (5.62%).  The other category results should be interpreted with caution, as there are only four observations.
* 90 Day Temperature: A one-degree (C) increase in 90-day water temperature prior to sampling is associated with a decrease in average seagrass cover by 0.88%.
* Mean Flood Turbidity/(Years Since Flood)$^2$: A one-unit increase in mean turbidity during the flood is associated with a decrease in average seagrass cover of 2.05%/(Years Since Flood)$^2$ (@fig-mb_turb_90d_fld).
* 90 Day Turbidity: A one-unit increase in 90-day turbidity (prior to sampling) is associated with a decrease in average seagrass cover by 7.18% (@fig-mb_turb_90d_fld).

```{r}
#| label: tbl-mb-covariates
#| tbl-cap: "ANOVA table of covariates for the categorical variables in the Moreton Bay water quality and geographic model."
#| results: asis
mb_tc_splm_mod %>%
  anova() %>%
  tidy() %>%
  filter(effects != "(Intercept)") %>%
  arrange(desc(statistic)) %>%
  mutate(effects = c("Tidal Range", "Sediment Type", "90 Day Temperature", "Mean Flood Turbidity/Years Since Flood^2", "90 Day Turbidity")) %>%
  filter(effects %in% c("Tidal Range", "Sediment Type")) %>%
rename(Effects = effects, df = df, `Chi-sq` = statistic, p.value = p.value) %>%
  xtable(digits = 3) %>%
  print(include.rownames = FALSE, floating = TRUE, booktabs = TRUE, comment = FALSE)
```

```{r}
#| label: fig-mb_turb_90d_fld
#| fig-cap: "Average total seagrass percent cover as a function of 90-day mean turbidity (left) and flood turbidity (right). Estimates (and confidence intervals) were evaluated at one year post flood, at the means of all other numeric variables, and for the intertidal range and sand sediment types."
pred_grid <- tibble(
  depth.range = "intertidal",
  sed.type = "Sand",
  temp.90d.mn = mean(dat$temp.90d.mn),
  turb.90d.mn = mean(dat$turb.90d.mn),
  turb.fld.mn = mean(dat$turb.fld.mn),
  yr.flood.recip2 = 1,
  xc = 75000, yc = 150000
) %>% 
  st_as_sf(crs = 3113, coords = c("xc", "yc"))

n_grid <- 200

pred_grid_turb.90d.mn <- pred_grid[rep(1, n_grid), ]
pred_grid_turb.90d.mn$turb.90d.mn <- seq(0, 2, length.out = n_grid)
pred_grid_turb.90d.mn[, c("fit", "lwr", "upr")] <- predict(mb_tc_splm_mod, newdata = pred_grid_turb.90d.mn, local = TRUE, interval = "confidence")
pred_grid_turb.fld.mn <- pred_grid[rep(1, n_grid), ]
pred_grid_turb.fld.mn$turb.fld.mn <- seq(0, 2, length.out = n_grid)
pred_grid_turb.fld.mn[, c("fit", "lwr", "upr")] <- predict(mb_tc_splm_mod, newdata = pred_grid_turb.fld.mn, local = TRUE, interval = "confidence")

p1 <- ggplot(pred_grid_turb.90d.mn, aes(x = turb.90d.mn)) +
  geom_line(aes(y = fit)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4, fill = "grey") +
  labs(x = "Turbidity (90 Day)", y = "Average Total Cover") +
  theme_bw(base_size = 14)


p2 <- ggplot(pred_grid_turb.fld.mn, aes(x = turb.fld.mn)) +
  geom_line(aes(y = fit)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4, fill = "grey") +
  labs(x = "Turbidity (Flood)", y = "Average Total Cover") +
  theme_bw(base_size = 14)

p1 + p2
```


We examined model diagnostics to better understand the overall model fit. First, we partitioned the model into different components of variability (@tbl-mb_varcomp) and found that 11.2% is attributable to the covariates (a quantity sometimes referred to as the Pseudo R-squared), 31.6% to the spatial random error, 7.3% to the EHMP subzone, and the remaining variability is independent random error. Second, we examined the decay behaviour of spatial autocorrelation. The estimated exponential spatial range ($\phi$) is 745.5 meters, which suggests two total seagrass cover observations are approximately (spatially) uncorrelated at a distance of $3\phi \approx 2$km, or two kilometers. Third, we analyzed the standardized model residuals, which have been decorrelated, to assess model assumptions and identify influential observations that had a sizable impact on model fit. Generally, the standardized residuals are centered around zero, though there is a slight right skew suggesting some observed total cover values are larger than expected (according to the model). Observations are referred to as "influential" if they highly influence model fit. Influence is often assessed using a Cook's distance threshold value of one [@cook1982residuals]. No observations exceed this threshold.

```{r}
#| label: tbl-mb_varcomp
#| tbl-cap: "Variance components in the Moreton Bay seagrass model."
#| results: asis
mb_tc_splm_mod %>%
  varcomp() %>%
  mutate(varcomp = c("Covariates", "Spatial Effect", "Independent Error", "EHMP Subzone")) %>%
  adorn_pct_formatting() %>%
  mutate(val = c(NA, coef(mb_tc_splm_mod, type = "spcov")[1:3])) %>%
  rename(`Variance Component` = varcomp, `Percent (of Total Variability)` = proportion, `Parameter Value` = val) %>%
  xtable(digits = 1) %>%
  print(include.rownames = FALSE, floating = TRUE, booktabs = TRUE, comment = FALSE)

```

## Block Kriging

A tremendous benefit of using the spatial linear model is that block Kriging [@ver2002sampling; @cressie2006block] can be used to predict the mean total seagrass cover bay-wide for any combination of covariates and geographic regions, even if these variables were not directly used in the model (see Appendix 1 for additional details). Block Kriging was undertaken, by financial year, for EHMP subzone, Marine Park zone, tidal range and for the whole of Moreton Bay.  


@fig-bk_mb_ehmp and @fig-bk_mb_zone show there are several EHMP subzones and Marine Park zones that have recovered to pre-flood levels (e.g., Deception Bay, Marine National Park Zone), while others have not (e.g., Southern Bay, General Use Zone). When we examine recovery by tidal range (@fig-bk_mb_tidal), we see that seagrass cover is recovering across all tidal ranges. When we examine recovery for the whole of Moreton Bay (@fig-bk_mb_all), the model suggests that seagrass cover has improved since the 2022 flood. 

```{r}
#| fig-width: 8
#| fig-height: 10
#| label: fig-bk_mb_ehmp
#| fig-cap: "Average total cover predictions by EHMP subzone. Red dots indicate the mean from the raw data. Black dots indicate the predicted mean in the region alongside a prediction interval."
bk_mb_ehmp <- readRDS(here("models", "bk_mb_ehmp.rds"))
ggplot(bk_mb_ehmp, aes(x = fy, y = fit)) +
  geom_point() +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.3) +
  geom_point(aes(x = fy, y = raw_mean), color = "red") +
  facet_wrap(~ ehmp.subzone, ncol = 2) +
  labs(x = "Financial Year", y = "Predicted Mean Total Cover") +
  theme_bw(base_size = 12)
```

```{r}
#| fig-width: 8
#| fig-height: 10
#| label: fig-bk_mb_zone
#| fig-cap: "Average total seagrass cover predictions by conservation zone. Red dots indicate the mean from the raw data. Black dots indicate the predicted mean in the region alongside a prediction interval."
bk_mb_zone <- readRDS(here("models", "bk_mb_zone.rds"))
ggplot(bk_mb_zone, aes(x = fy, y = fit)) +
  geom_point() +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.3) +
  geom_point(aes(x = fy, y = raw_mean), color = "red") +
  facet_wrap(~ zone.type, ncol = 2) +
  labs(x = "Financial Year", y = "Predicted Mean Seagrass Cover") +
  theme_bw(base_size = 12)
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| label: fig-bk_mb_tidal
#| fig-cap: "Average total seagrass cover predictions by tidal range. Red dots indicate the mean from the raw data. Black dots indicate the predicted mean in the region alongside a prediction interval."
bk_mb_tidal <- readRDS(here("models", "bk_mb_tidal.rds"))
ggplot(bk_mb_tidal, aes(x = fy, y = fit)) +
  geom_point() +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.3) +
  geom_point(aes(x = fy, y = raw_mean), color = "red") +
  facet_wrap(~ depth.range, ncol = 2) +
  labs(x = "Financial Year", y = "Predicted Mean Seagrass Cover") +
  theme_bw(base_size = 12)
```

```{r}
#| fig-width: 4
#| fig-height: 3
#| label: fig-bk_mb_all
#| fig-cap: "Average total seagrass cover predictions in Moreton Bay. Red dots indicate the mean from the raw data. Black dots indicate the predicted mean in the region alongside a prediction interval."
bk_mb_all <- readRDS(here("models", "bk_mb_all.rds"))
ggplot(bk_mb_all, aes(x = fy, y = fit)) +
  geom_point() +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.3) +
  geom_point(aes(x = fy, y = raw_mean), color = "red") +
  labs(x = "Financial Year", y = "Predicted Mean Seagrass Cover") +
  theme_bw(base_size = 12)
```

\clearpage

# Resilience Model {#sec-mb_resilience}

The seagrass cover model quantifies bay-wide effects of influential drivers on seagrass cover and enables assessment of recovery trajectories at management-relevant scales. However, it does not quantify site-specific resilience. @gilby2023drivers propose identifying hotspots (i.e., brightspots, sites performing better than expected) and coldspots (sites performing worse than expected) based on model residuals exceeding confidence interval bounds. They argue that hotspots and coldspots offer significant utility for understanding drivers of site-specific resilience and informing management decisions. @gilby2023drivers primarily focused on specific hotspots, linking environmental variables to those sites. We build on that work by using the seagrass model residuals as a proxy for site-specific resilience, and then characterizing the most important drivers of that resilience.

## Data

We used the raw seagrass model residuals (@fig-resid_fy_mb) as a the response variable in a second spatial linear model used to study site-specific resilience in Moreton Bay. Model covariates included variables describing seagrass richness and dominant species, and bathymetry, as well as distance to the nearest coral, seagrass, mangrove, or gastropod habitat, and the total area in msq of coral, seagrass, mangrove, or gastropod habitat within 500 m of the observation point (@tbl-res_vars_mb). EHMP subzone was again included as a random effect. Because we were interested in hypothesis testing rather than parsimony, we included all potential covariates as fixed effects in the model. Potential covariates have the following structure:

* Coral Habitat Total Area is numeric from 0 msq to 616,180 meters squared (msq).
* Seagrass Habitat Total Areais numeric from 0 msq to 773,161 msq.
* Mangrove Habitat Total Area is numeric from 0 msq to 561,317 msq.
* Gastropod Habitat Total Area is numeric from 0 msq to 114,795 msq.
* Distance to Coral Habitat is numeric from 0 m to 35,387 m.
* Distance to Seagrass Habitat is numeric from 0 m to 11,382 m.
* Distance to Mangrove Habitat is numeric from 0 m to 27,617 m.
* Distance to Gastropod Habitat is numeric from 0 m to 32,034 m.
* Habitat Count is numeric (integer) from 0 to 4 unique habitats (coral, seagrass, mangrove, gastropod).
* Species Richness is numeric (integer) from 0 to 4 unique species (possible species: "Absent", "CS", "HD", "HO", "HS", "SI", and "ZMHU").
* Dominant Species represents the dominant seagrass species and has seven levels: "Absent", "CS", "HD", "HO", "HS", "SI", and "ZMHU".
* Bathymetry Mean Depth is the GBR mean depth within 500 m of the seagrass observation point, excluding terrestrial areas, and is numeric from -35.95 m to 0.79 m.
* Bathymetry Mean Curvature  is the GBR mean curvature within 500 m of the seagrass observation point, excluding terrestrial areas, and is numeric from -0.002 degrees to 0.0001 degrees.
* Bathymetry Mean Slope is the GBR mean slope within 500 m of the seagrass observation point, excluding terrestrial areas, and is numeric from 0.009 degrees to 4.27 degrees.
* EHMP subzone has eleven categorical levels: Bramble Bay, Broadwater, Central Bay, Deception Bay, Eastern Banks, Moreton Bay Open Coastal Waters, North-Eastern Moreton Bay, Northern Banks Open Coastal Waters, South-Eastern Moreton Bay, Southern Bay, and Waterloo Bay (@fig-ehmp_subzone). 

We expected there to be nonlinear relationships between the response and covariates and we accounted for them using B-spline basis functions [@deboor1978practical; @eilers1996flexible]. The degree of these functions was informed by partial dependence plots derived from a random forest model fitted to the seagrass model residuals (@tbl-res_vars_mb). 
```{r}
#| fig-width: 8
#| fig-height: 8
#| label: fig-resid_fy_mb
#| fig-cap: "Seagrass model residuals for resilience analysis in Moreton Bay, by financial year."
dat$residuals <- residuals(mb_tc_splm_mod)
dat2 <- dat %>%
  arrange(residuals, fy)
ggplot(dat2) +
  geom_sf(data = bay, fill = "white") +
  geom_sf(data = dat2, aes(color = residuals), size = 0.8) +
  scale_color_viridis_c(name = "Residuals", option = "H") +
  scale_x_continuous(breaks = c(153.1, 153.3, 153.5)) +
  facet_wrap(~ fy) +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```


```{r}
#| label: tbl-res_vars_mb
#| tbl-cap: "Covariates included in the resilience model, including whether they were a fixed or random effect, and the nonlinearity degree of the B-splines basis function."
#| results: asis
res_vars <- tibble(
  Variable = c("Coral Habitat Total Area", "Seagrass Habitat Total Area",
              "Mangrove Habitat Total Area", "Gastropod Habitat Total Area",
              "Distance to Coral Habitat", "Distance to Seagrass Habitat",
              "Distance to Mangrove Habitat", "Distance to Gastropod Habitat",
              "Species Richness", "Habitat Count", "Dominant Species",
              "Bathymetry Mean Depth", "Bathymetry Mean Curvature", "Bathymetry Mean Slope",
              "EHMP Subzone"), 
  Effect = c(rep("Fixed", 14), "Random"),
  `Degree` = c(rep(2, 4), 3, 2, 2, 2, 1, 2, "NA", 2, 2, 2, "NA")
)
res_vars %>%
  xtable() %>%
  print(include.rownames = FALSE, floating = TRUE, booktabs = TRUE, comment = FALSE)

```

## Statistical Modeling

```{r}
#| label: tbl-mb-res-fixed
#| tbl-cap: "Fixed effects table of covariates in the Moreton Bay resilience model."
#| results: asis
mb_tc_splm_mod_res <- readRDS(here("models", "mb_tc_splm_mod_res.rds"))
mb_tc_splm_mod_res %>%
  tidy() %>%
  mutate(term = c("Intercept (Absent)", "CS", "HD", "HO", "HS", "SI", "ZMHU",
                  "Species Richness", "Coral Area Deg1", "Coral Area Deg2",
                  "Seagrass Area Deg 1", "Seagrass Area Deg1", "Mangrove Area Deg1",
                  "Mangrove Area Deg2",
                  "Gastropod Area Deg1", "Gastropod Area Deg2", 
                  "Habitat Count Deg1", "Habitat Count Deg2", 
                  "Coral Distance Deg1", "Coral Distance Deg2", "Coral Distance Deg3",
                  "Seagrass Distance Deg1", "Seagrass Distance Deg2",
                  "Mangrove Distance Deg1", "Mangrove Distance Deg2",
                  "Gastropod Distance Deg1", "Gastropod Distance Deg2",
                  "Bathymetry Mean Depth Deg1", "Bathymetry Mean Depth Deg2",
                  "Bathymetry Mean Slope Deg1", "Bathymetry Mean Slope Deg2",
                  "Bathymetry Mean Curvature Deg1", "Bathymetry Mean Curvature Deg2")) %>%
  select(-std.error) %>%
  rename(Term = term, Estimate = estimate, z = statistic, p.value = p.value) %>%
  xtable(digits = 3) %>%
  print(include.rownames = FALSE, floating = TRUE, booktabs = TRUE, comment = FALSE)
```


@tbl-mb-res-fixed shows the fixed effects table from the fitted model. The reference group is the absent dominant species, and the remaining dominant species terms represent deviations from this baseline. Species richness, the only numeric variable with a single level, had a large test statistic and small $p$-value ($p$-value $<$ 0.01). The other numeric variables were modeled using a B-spline basis of at least degree two to capture nonlinearities. While we can assess the overall significance of species richness in @tbl-mb-res-fixed, we must perform an analysis of variance (ANOVA) to assess the overall significance the remaining variables with multiple terms. @tbl-mb-res-anova shows these ANOVA results.

```{r}
#| label: tbl-mb-res-anova
#| tbl-cap: "ANOVA table of covariates in the Moreton Bay resilience model."
#| results: asis
mb_tc_splm_mod_res %>%
  anova() %>%
  tidy() %>%
  filter(effects != "(Intercept)") %>%
  arrange(p.value) %>%
  mutate(effects = c("Dominant Species", "Species Richness", "Bathymetry Mean Depth",
         "Bathymetry Mean Curvature", "Gastropod Habitat Total Area", 
         "Bathymetry Mean Slope", "Coral Habitat Total Area",
         "Seagrass Habitat Total Area", "Distance to Mangrove Habitat", 
         "Distance to Seagrass Habitat", "Habitat Count", 
         "Distance to Gastropod Habitat", "Distance to Mangrove Habitat", 
         "Distance to Coral Habitat")) %>%
  filter(!effects %in% c("Species Richness")) %>%
  rename(Effects = effects, df = df, `Chi-sq` = statistic, p.value = p.value) %>%
  xtable(digits = 3) %>%
  print(include.rownames = FALSE, floating = TRUE, booktabs = TRUE, comment = FALSE)
```

We found strong evidence ($p$-value < 0.001) that dominant species, species richness, and bathymetry mean depth are associated with resilience: 

* Dominant Species: The estimated (marginal) average resilience score is largest for CS (38.4), followed by ZMHU (18.8), HS (10.8), SI (10.7), HO (3.7), and HD (-3.1).
* Species Richness: An increase of species richness by one species was associated with an average increase in resilience by approximately two units (@fig-mb_res_rich_bath).
* Bathymetry Mean Depth: Deeper sites tended to be more resilient than shallower sites, on average (@fig-mb_res_rich_bath). 

We found moderate evidence ($p$-value < 0.1) that bathymetry mean curvature, gastropod habitat total area, and bathymetry mean slope are associated with resilience: 

* Bathymetry Mean Curvature: Sites with more curvature tended to be less resilient than sites with less curvature.
* Gastropod Habitat Total Area: Sites with more gastropod habitat area tended to be more resilient than sites with less gastropod habitat area.
* Bathymetry Mean Slope: Sites with less extreme slopes (between zero and one) tended to have increasing resilience with slope, while sites with more extreme slopes (greater than one) tended to have decreasing resilience with slope.

We found little evidence ($p$-value > 0.1) the remaining variables were associated with resilience, after controlling for other model covariates.


```{r}
#| label: fig-mb_res_rich_bath
#| fig-cap: "Average species richness (left) and bathymetry mean depth (right) effects on resilience. Estimates (and confidence intervals) were evaluated at the means of all other numeric variables and for the ZMHU species in the Southern Bay."
#| out-width: "80%"
pred_grid <- tibble(
  dom.spp = "ZMHU",
  sg.rich = mean(dat$sg.rich),
  coral.msq = mean(dat$coral.msq),
  seagr.msq = mean(dat$seagr.msq),
  u.mang.msq = mean(dat$u.mang.msq),
  shell.msq = mean(dat$shell.msq),
  hab.count = mean(dat$hab.count),
  coral.dist.m = mean(dat$coral.dist.m),
  seag.dist.m = mean(dat$seag.dist.m),
  u.mg.dist = mean(dat$u.mg.dist),
  shell.dist.m = mean(dat$shell.dist.m),
  bath.m.mean = mean(dat$bath.m.mean),
  rslope.dg.mean = mean(dat$rslope.dg.mean),
  rcurv.dg.mean = mean(dat$rcurv.dg.mean),
  ehmp.subzone = "Southern Bay",
  xc = 75000, yc = 150000
) %>% 
  st_as_sf(crs = 3113, coords = c("xc", "yc"))

n_grid <- 200

pred_grid_sg.rich <- pred_grid[rep(1, n_grid), ]
pred_grid_sg.rich$sg.rich <- seq(0, 4, length.out = n_grid)
pred_grid_sg.rich[, c("fit", "lwr", "upr")] <- predict(mb_tc_splm_mod_res, newdata = pred_grid_sg.rich, local = TRUE, interval = "confidence")

pred_grid_bath.m.mean <- pred_grid[rep(1, n_grid), ]
pred_grid_bath.m.mean$bath.m.mean <- seq(quantile(dat$bath.m.mean, 0.01), quantile(dat$bath.m.mean, 0.99), length.out = n_grid)
pred_grid_bath.m.mean[, c("fit", "lwr", "upr")] <- predict(mb_tc_splm_mod_res, newdata = pred_grid_bath.m.mean, local = TRUE, interval = "confidence")

p1 <- ggplot(pred_grid_sg.rich, aes(x = sg.rich)) +
  geom_line(aes(y = fit)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4, fill = "grey") +
  labs(x = "Species Richness", y = "Average Resilience") +
  theme_bw(base_size = 14)


p2 <- ggplot(pred_grid_bath.m.mean, aes(x = bath.m.mean)) +
  geom_line(aes(y = fit)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4, fill = "grey") +
  labs(x = "Bathymetry Mean Depth", y = "Average Resilience") +
  theme_bw(base_size = 14)

p1 + p2
```

As mentioned previously, we also fit a random forest model to the seagrass model residuals. Largely, the results were similar to the spatial linear model. Notably, for the random forest:

* Dominant species was most important, with a similar ranking of resilient species as for the spatial linear model approach (e.g., CS, ZMHU most resilient and Absent least resilient).
* Species richness was second-most important, with a general increasing trend in resilience with richness.
* Distance to gastropod habitat was third-most important, with a decreasing trend at relatively short distances (< 15,000 m) and an increasing trend at large distances (> 15,000 m).
* Bathymetry mean depth was the fourth-most important variable, with a general decreasing trend with decreasing depth (@fig-rf_bathy), which is similar to the spatial linear model results.
* Seagrass area, distance to mangrove habitat, and distance to coral habitat had similar importance as the remaining bathymetry variables.
* The remaining variables (distance to seagrass habitat, mangrove area, EHMP subzone, habitat count, coral area, and gastropod area) were least important.

The consistency between the two resilience models and their variable importance rankings reinforce the spatial linear model's identification of dominant species, species richness, and bathymetry mean depth as key resilience drivers.

```{r}
#| label: fig-rf_bathy
#| fig-cap: "Random forest partial dependence for resilience as a function of bathymetry mean depth."
#| out-width: "50%"
mb_tc_rf_mod_res <- readRDS(file = here("models", "mb_tc_rf_mod_res.rds"))
pred.grid <- tibble(bath.m.mean = seq(quantile(dat$bath.m.mean, 0.01), quantile(dat$bath.m.mean, 0.99), length.out = 6))
pd <- partial(mb_tc_rf_mod_res, pred.var = "bath.m.mean", pred.grid = pred.grid) 
p1 <- ggplot(pd, aes(x = bath.m.mean, y = yhat)) +
  geom_line(linewidth = 2) +
  labs(x = "Bathymetry Mean Depth", y = "Average Resilience") +
  theme_bw(base_size = 14)
p1
```

# Limitations

While the spatial linear model was useful for characterizing broad patterns in seagrass cover related to tidal range, sediment type, temperature, and turbidity and site-specific patterns capturing resilience, all while accounting for spatial dependence, there are some limitations. First, the eReefs water-quality covariates are based on modeled data that are not directly observed in the field, potentially propagating modeling error. Second, there are many correlated water-quality variables, making it difficult for any model to isolate the effects of each component of water quality pre- and post-flood. Third, the lack of a covariate describing mean salinity in the 90 days prior to sampling makes it challenging to directly study the effect of mean salinity during the flood period. Fourth, seagrass cover data was not collected in the first four months following the flood. It is therefore unclear whether the seagrass data collected in the 2022-2023 financial year fully describe flood impacts. Fifth, covariates representing local-scale drivers of seagrass cover (e.g., light attenuation, biological interactions, etc.) were not available, which might explain some additional variability in the response through space and time. Sixth and finally, the location of sample sites was not generated randomly and could be subject to conscious or unconscious selection bias.

# Revisiting the Research Questions

**Research Question 1**: Can we identify water quality covariates that act as triggers for seagrass cover management responses? Are these triggers different based on whether they result from a specific event (e.g., flood) or are ambient?

We identified ambient 90-day turbidity, turbidity during the flood event, and ambient 90-day water temperature as significant water quality covariates in the seagrass cover model. Our results suggest that ambient and flood-related turbidity act through distinct mechanisms and at different magnitudes; a one-unit increase in 90-day mean turbidity was associated with a 7.18 percentage point decrease in total cover, while the effect of flood turbidity decayed with time since the flood. Across tidal ranges, intertidal sites supported the highest average total cover (26.89%), followed by shallow subtidal (16.51%) and deep subtidal (4.88%), suggesting that management thresholds may need to be calibrated differently depending on depth zone. Similarly, sediment type affected baseline cover levels, as sites with muddy sand and sandy mud substrates supported higher average seagrass cover than sand or rock sites. Together, these results suggest that management triggers should differ depending on whether the driver is chronic ambient turbidity or an acute flood event, and should further account for the tidal and sediment context of the sites in question.


**Research Question 2**: Did seagrass cover improve after the flood event(s)? If so, how and where?

Total seagrass cover improved throughout Moreton Bay between 2022–2023 and 2023–2024, suggesting the system has been recovering since the flood. However, recovery was spatially heterogeneous; Deception Bay and the Marine National Park zone returned to pre-flood cover levels, while Southern Bay and the General Use Zone lagged considerably behind. Recovery was also relatively slow compared to the magnitude of the impact in the intertidal tidal range compared to shallow and deep subtidal ranges, suggesting that intertidal sites may warrant prioritized management attention immediately following floods.

**Research Question 3**: Can we identify site-specific resilience indicators, which describe where recovery is likely to be enhanced or impeded?  What factors contribute to seagrass resilience?

We identified several site-specific indicators of seagrass resilience. Dominant species, species richness, and bathymetry mean depth were the strongest predictors of seagrass cover resilience, with sites dominated by CS or ZMHU species, higher species richness, and greater water depth tending to outperform model expectations and hence, infer greater resilience. Gastropod habitat area, bathymetric curvature, and slope provided moderate additional explanatory power. These findings may suggest that deeper, species-rich meadows dominated by CS and ZMHU exhibit a high level of resilience, while sites with low species richness and/or dominated by other species may require more active intervention to support recovery.

# Summary

We applied a rigorous, spatially explicit statistical framework to characterize the drivers, recovery, and resilience of seagrass cover in Moreton Bay across a three-year monitoring period that included the 2022 Moreton Bay flood event. The analytical approach taken here is well-suited to the structure and scale of the data, and the findings have direct implications for adaptive management of the bay's seagrass ecosystem.

The use of spatial linear models as the primary analytical framework is well-justified from both statistical and ecological perspectives. Total percent seagrass cover observations collected across the bay are inherently spatially structured, as nearby sites share similar environmental conditions, water quality and disturbance histories, as well as ecological communities. Models that ignore this dependence may well provide misleading results and inference. By explicitly modeling spatial dependence via a spatial covariance function, the analyses presented here produce more reliable estimates of covariate effects and more honest characterizations of uncertainty than a conventional non-spatial model provides. These models are typically also more interpretable and characterize uncertainty more effectively than machine learning models. The inclusion of EHMP subzone as a random effect further accounts for structured variation among management-relevant regions of the bay, ensuring that covariate estimates reflect within-subzone patterns rather than being confounded by between-subzone differences in baseline cover. The block Kriging helps understand bay-wide trends and can immediately characterize bay-wide predictions on the original total seagrass cover scale (0-100), rather than a transformed scale required for generalized linear models; providing a flexible framework for scenario modeling under future flood conditions and climate scenarios.

The use of a two-stage modeling framework -- first modeling total seagrass cover as a function of water quality and geographic drivers, then modeling the seagrass model residuals as a function of site-level resilience covariates -- is a structured approach to separating bay-wide environmental patterns from site-specific ecological resilience capacity. This builds directly on the framework proposed by @gilby2023drivers to help understand drivers of average resilience and provide useful insight needed to inform broader conservation strategies. The robustness of the findings is further supported by the general consistency of results across multiple modeling approaches (e.g., spatial beta regression model, random forest, etc.), which provides evidence that the relationships identified are genuine rather than resulting solely from specific modeling frameworks. 

Overall, the results described in this report provide a quantitative, spatially structured foundation for adaptive decision making. Together, the identification of important water quality and geographic variables, bay-wide predictions via block Kriging, and the characterization of site-level resilience indicators provide data-driven evidence that can inform temporally targeted, context-specific interventions throughout Moreton Bay.

\clearpage

# References {.unnumbered}

::: {#refs}
:::

# Acknowledgements
The seagrass cover dataset was provided by Science Under Sail Australia (SUSA; https://www.scienceundersail.com.au/). The eReefs datasets (model simulations, satellite or in-situ observations) and software were produced as part of the eReefs project (ereefs.org.au), which is a collaboration between Australia’s national science agency CSIRO, the Australian Institute of Marine Science (AIMS) and the Queensland Government, with observations obtained through the Integrated Marine Observing System (IMOS) and the Great Barrier Reef Marine Park Authority (GBRMPA). eReefs is funded by the Australian Government’s Reef Trust.

\clearpage

# Appendix 1. Spatial Linear Models and Random Forest Models {.unnumbered}

The spatially explicit statistical models (i.e., spatial models) we used to answer these research questions are related to commonly used linear and generalized linear models but incorporate _spatial dependence_. Spatial dependence measures how similar two sites are based on their distance. Formally, we can build spatial dependence directly into spatial models by leveraging Tobler's First Law of Geography [@tobler1970computer], which states succinctly that "nearby sites tend to be more similar than distant sites". Not only is this intuitive from a "first principles" perspective, but spatial models also provide many advantages for modeling spatial data compared to models that assume independence among observations (i.e., nonspatial models). First, nonspatial models tend to underestimate uncertainty, yielding covariate standard errors and p-values for covariates that are too small and corresponding confidence intervals that are too narrow. Second, spatial models improve prediction at unsampled locations by borrowing strength from nearby observations, yielding more precise predictions and corresponding standard errors that vary by location. Third, spatial models decompose model error into separate spatial and nonspatial components, informing residual and resilience structure in ways that nonspatial models cannot. Fourth, spatial models can provide ecologically meaningful insight into the spatial dependence structure of the process, informing relationships among measured and unmeasured variables as well as future sampling plans. @zimmerman2024spatial provide a thorough review of spatial models for ecological and environmental data, providing more detail and justification for the claims made above.

Spatial models also offer many benefits over commonly used machine learning algorithms like random forests [@breiman2001random; @cutler2007random; @james2017introduction; @kuhn2022tidy]. First, traditional machine learning algorithms do not incorporate spatial dependence and, by consequence, often perform worse than spatial models at characterizing important covariates and predicting at unobserved locations [@fox2020comparing]. Second, machine learning algorithms struggle to quantify uncertainty in a principled manner, something spatial models excel at. Third, covariate inference is more straightforward using spatial models, as they naturally provide p-values that enable structured testing of hypotheses. Fourth, spatial models provide a rich set of diagnostics like leverage and Cook's distances [@montgomery2021introduction], quantities not generally defined for machine learning algorithms. This is not to say that machine learning algorithms cannot be very useful in ecological data analyses, but rather that spatial models are sometimes a more effective tool for modeling spatial data. 

Though there are many benefits of spatial models over machine learning algorithms, the reverse is also true. It is much simpler to capture complex nonlinearities using "out-of-the-box" machine learning algorithms, bypassing the tuning often required for spatial models that involve nonlinear components like splines. This can be especially useful for ecological data, which often exhibit nonlinear relationships [@fox2017assessing]. Machine learning algorithms also typically rely on few, if any, distributional assumptions and can be more robust against unusual observations like outliers [@hastie2009elements]. Machine learning algorithms also tend to be much more computational efficient than spatial models. Importantly, machine learning algorithms and spatial models can be complementary tools that help shed insight into complicated ecological processes (something we reiterate in the upcoming analyses). An exciting and developing area of statistical research involves incorporating spatial modeling strategies directly into machine learning algorithms [@saha2023random; @heaton2025scalable].

The first spatially explicit model we define is the spatial linear model. The spatial linear model is a _mixed-effects model_ [@pinheiro2000mixed; @bolker2015linear] written as

$$
y_i = \beta_0 + \beta_1x_{1, i} + \beta_2x_{2, i} + ... + \beta_p x_{p, i} + \tau_i + \epsilon_i,
$$ {#eq-splm1}
where $i = 1, 2, ..., n$ indexes the $n$ observations. The quantity $y_i$ is the response variable (for observation $i$). The quantity $\beta_j$ is a slope parameter that controls the average effect of the covariate $x_{j, i}$ on $y_i$ (while holding all other covariates constant). The random error $\tau_i$ is a spatial random error, and the random error $\epsilon_i$ is the nonspatial (i.e., independent) random error. The spatial random error is spatially structured, capturing spatial residual effects like environmental gradients we do not have data to fully represent. The nonspatial random error is not spatially structured and captures nonspatial residual effects like measurement error and microscale variation. The difference between a traditional (i.e., nonspatial) linear model and a spatial linear model is the inclusion of $\tau_i$ (in the spatial linear model). The spatial random error implied by $\tau_i$ is governed by a _spatial covariance function_. The spatial covariance function allows the model to incorporate spatial dependence, yielding the improved model performance previously described. @fig-spcov_fns shows three different spatial covariance functions that describe different ways the spatial covariance decays with distance. Model parameters and standard errors are estimated from data using likelihood-based [@patterson1971recovery; @harville1977maximum; @wolfinger1994computing] or semivariogram-based [@cressie1985fitting; @curriero1999composite] estimation methods. Often, the spatial linear model in @eq-splm1 is written more compactly using _matrix notation_:
$$
\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
$$ {#eq-splm2}
where $\mathbf{X}$ is a matrix that holds the $j$ covariates for each of $n$ observations, and $\mathbf{y}$, $\boldsymbol{\beta}$, $\boldsymbol{\tau}$, and $\boldsymbol{\epsilon}$ are vectors that contain each of the $n$ or $j$ quantities from @eq-splm1. The spatial linear model can be adjusted to account for nonspatial random effects just as one would add random effects to a nonspatial linear mixed model:
$$
\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \mathbf{Z}\mathbf{u} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
$$ {#eq-splm3}
where $\mathbf{Z}$ is a random effect design matrix (dimension $n \times k$) that indexes each of the $k$ random effects in $\mathbf{u}$. For more information about specific forms of various spatial covariance functions, see @schabenberger2017statistical and @zimmerman2024spatial.

```{r}
#| label: fig-spcov_fns
#| fig-cap: "Three spatial covariance functions that represent different spatial dependence behavior."
#| out-width: "75%"
range <- 50
spcor_dat <- tibble(
  h = seq(0, range*2),
  Exponential = exp(-h/range),
  Gaussian = exp(-h^2/range^2),
  Spherical = (1 - 1.5*h/range + 0.5*h^3/range^3)*(h <= range)
) %>%
  rename(Distance = h) %>%
  pivot_longer(cols = c(-Distance), names_to = "Function", values_to = "Spatial Covariance")
ggplot(spcor_dat, aes(x = Distance, y = `Spatial Covariance`, color = Function, linetype = Function)) +
  geom_line(linewidth = 2) +
  theme_bw(base_size = 14)
```

Like the traditional linear model, the spatial linear model can capture complex nonlinearities in the data through the use of interaction effects [@faraway2002practical; @lenth2025emmeans], splines [@chambers1992statistical; @ruppert2003semiparametric], additive models [@wood2017generalized; @pedersen2019hierarchical], and penalized regression models like the ridge and lasso [@tibshirani1996regression]. Generally, this is accomplished by leveraging _basis functions_ that break down nonlinear behavior into a set of additive linear components that may include penalty terms which guard against overfitting. Furthermore, there is no assumption that $y_i$ itself is normally distributed; rather, distributional assumptions are placed on the random errors after incorporating the slope parameters. This clarification implies the spatial (and nonspatial) linear models can be quite effective at modeling a range of data that are not normally distributed. For highly nonnormal data, tools like generalized linear models, which we discuss later, can be helpful. Finally, under general conditions, the slope parameters of the spatial (and nonspatial) linear models are approximately normally distributed with a large sample size (see @casella2024statistical for a general justification via the Central Limit Theorem and @zhang2005towards for nuances regarding spatial data). This is beneficial because it justifies valid hypothesis testing on the slope coefficients (often of primary ecological interest), no matter the distribution of the data (given certain assumptions are satisfied).

One of the most useful applications of the spatial linear model is predicting the response variable at unobserved locations. A very useful spatial prediction method called Kriging has a rich theoretical justification [@cressie1990origins; @stein1999interpolation]. Kriging is the spatial analogue to best linear unbiased prediction (BLUP) for mixed models [@henderson1975best]. Via Kriging, the spatial linear model produces BLUPs at each unobserved location of interest, alongside prediction intervals to characterize uncertainty. While useful, sometimes the goal is to create a BLUP for an entire region of interest based on realized (i.e., observed) covariate values. This technique, which builds upon the theoretical foundations of Kriging, is called block Kriging [@ver2002sampling; @cressie2006block]. The intuition is to make point predictions at a very dense grid in the study region and then average these predictions, appropriately characterizing uncertainty. Together, Kriging and block Kriging provide a complete set of tools for spatial prediction, given the fit of a spatial linear model.

Spatial generalized linear models are an extension of spatial linear models to highly nonnormal data [@christensen2002bayesian; @zhang2002estimation; @diggle2007model; @bonat2016practical]. @verhoef2024marginal proposed a novel application of the Laplace approximation to enable restricted maximum likelihood estimation of spatial generalized linear models. This model formulation builds upon @eq-splm1 and is given by
$$
g(\mu_i) = \beta_0 + \beta_1x_{1, i} + \beta_2x_{2, i} + ... + \beta_p x_{p, i} + \tau_i + \epsilon_i,
$$ {#eq-spglm1}
where $g(\mu_i)$ is a "link function" that links the distribution of $y_i$ to a linear function of its mean, $\mu_i$. Common link functions include the log link for count and skewed data and the logit link for binomial and proportion data (@tbl-glm-families). Using the same argument as for spatial linear models, nonspatial random effects can be added to spatial generalized linear models.

```{r}
#| label: tbl-glm-families
#| tbl-cap: "Common generalized linear model families, link functions, and appropriate data types."
#| results: asis
glm_table <- data.frame(
  Family = c("Binomial", "Poisson", "Negative Binomial",
             "Beta", "Gamma", "Inverse Gaussian"),
  Link_Function = c("$f(\\mu) = \\log(\\mu / (1 - \\mu))$",
                    "$f(\\mu) = \\log(\\mu)$",
                    "$f(\\mu) = \\log(\\mu)$",
                    "$f(\\mu) = \\log(\\mu / (1 - \\mu))$",
                    "$f(\\mu) = \\log(\\mu)$",
                    "$f(\\mu) = \\log(\\mu)$"),
  Link_Name = c("Logit", "Log", "Log", "Logit", "Log", "Log"),
  Data_Type = c("Binary; Binary Count", "Count", "Count",
                "Proportion", "Skewed", "Skewed")
) %>%
  rename(Family = Family, `Link Function` = Link_Function,
         `Link Name` = Link_Name, `Data Type` = Data_Type) %>%
  xtable() %>%
  print(include.rownames = FALSE, floating = TRUE, booktabs = TRUE, comment = FALSE,
        sanitize.text.function = identity)
```

Spatial models are complicated, both theoretically and computationally. This has contributed to their relative underuse in the ecological community. However, recent advances in open-source software have made spatial models more accessible to ecologists who have strong subject matter expertise but may lack sufficient statistical background or computational skills to implement them from scratch. One such advancement is the `spmodel` **R** package [@dumelle2023spmodel], which emulates base-**R** functions like `lm()` and `glm()` to account for spatial dependence via `splm()` and `spglm()`, respectively. Familiar **R** functions like `summary()` and `predict()` can be used directly on models fit via `spmodel`. Importantly, `spmodel` also has tools for spatial random forest residual modeling [@fox2020comparing], applications to large data of several thousand observations via spatial indexing [@ver2023indexing], and cross validation [@stone1974cross; @roberts2017cross].

```{r}
#| label: fig-time_since_flood
#| fig-cap: "Time since the flood against the squared reciprocal of time since the flood."
#| eval: false
# dat_yr.flood <- dat %>%
#   filter(yr.flood > 0) %>%
#   mutate(yr.flood = seq(min(.$yr.flood), max(.$yr.flood), length.out = NROW(.))) %>%
#   mutate(yr.flood.recip2 = 1/yr.flood^2)
# ggplot(dat_yr.flood, aes(x = yr.flood, y = yr.flood.recip2)) +
#   geom_line() +
#   labs(x = "Years Since Flood", y = expression("1/(Years Since Flood)"^2)) +
#   theme_bw(base_size = 12)
```

```{r}
#| label: fig-cover_tidal_mb
#| fig-cap: "Total seagrass cover in Moreton Bay by tidal class and financial year. Boxplot width is proportional to the number of observations in each x-axis level (within financial year)."
#| eval: false
# dat_tidal <- dat %>%
#   mutate(depth.range = case_when(
#     depth.range == "deep subtidal" ~ "DSt",
#     depth.range == "shallow subtidal" ~ "SSt",
#     depth.range == "intertidal" ~ "Int",    
#   )) %>%
#   mutate(depth.range = factor(depth.range, levels = c("Int", "SSt", "DSt")))
# ggplot(dat_tidal, aes(x = depth.range, y = total.cover)) +
#   geom_boxplot(varwidth = TRUE, size = 0.5) +
#   facet_wrap(~ fy) + 
#   labs(x = "Tidal Class", y = "Total Cover") +
#   coord_cartesian(ylim = c(0, 100)) +
#   theme_bw(base_size = 14) +
#   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

```{r}
#| label: fig-cover_turb_mb
#| fig-cap: "Total seagrass cover in Moreton Bay by turbidity class (whether exceeding 0.5 NTU) and financial year. Generally, increased turbidity is associated with lower median total cover. Boxplot width is proportional to the number of observations in each x-axis level (within financial year). The y-axis is truncated at 50, though a few observations have higher total seagrass cover."
#| eval: false
# dat_turb <- dat %>%
#   mutate(exceed_0.5 = factor(if_else(turb.90d.mn > 0.5, "Yes", "No"), levels = c("Yes", "No")))
# ggplot(dat_turb, aes(x = exceed_0.5, y = total.cover)) +
#   geom_boxplot(varwidth = TRUE) +
#   facet_wrap(~ fy) + 
#   labs(x = "Exceeds 0.5 NTU", y = "Total Cover") +
#   coord_cartesian(ylim = c(0, 100)) +
#   theme_bw(base_size = 14)
```


```{r}
#| eval: false
# mb_tc_rf_mod_res <- readRDS(here("models", "mb_tc_rf_mod_res.rds"))
# pred_grid_rf <- pred_grid %>%
#   st_drop_geometry() %>%
#   select(bath.m.mean) %>%
#   mutate(dom.spp = "ZMHU", ehmp.subzone = "Southern Bay") %>%
#   slice(seq(1, n(), by = 10))
# pred_grid_rf <- partial(mb_tc_rf_mod_res, pred.var = "bath.m.mean", pred.grid = pred_grid_rf)
```
