---
title: "DETSI Report"
author: "EP Consulting and Dumelle Consulting"
format:
  html:
    toc: true
execute: 
  echo: false
  message: false
  warning: false
  error: true
embed-resources: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r}
# loading software
library(tidyverse)
library(here)
library(ranger)
library(sf)
library(knitr)
library(broom)
library(spmodel)
library(pdp)
library(emmeans)
library(janitor)
```


# Background

Accurately characterizing drivers and resilience of seagrass cover is essential to improving the ecological health of the Moreton Bay and Wide Bay ecosystems. Over the past few decades, various scientific efforts have informed our current understanding of trends, behaviors, and flood responses of seagrass cover in both Moreton Bay and Wide Bay [@preen1995flood; @mcmahon2005herbicide; @york2022post] [@abal1996seagrass; @udy1997growth; @samper2016organic; @ovsyanikova2026spatial] as well as Australia more widely [@kilminster2015unravelling; @gilby2018seagrass; @o2018seagrass]. Remote sensing has offered new capabilities for seagrass cover mapping [@roelfsema2009integrated], while innovative statistical approaches have offered novel methods for studying resilience at site-specific scales [@gilby2023drivers]. Our work contributes to this body of literature by applying spatially explicit statistical models [@cressie1993statistics] to study seagrass cover drivers, resilience, and recovery in both Moreton Bay and Wide Bay. Specifically, we provide and contextualize answers to four specific research questions, applicable to both Moreton Bay and Wide Bay:

1. Can we identify water quality covariates that act as triggers for seagrass cover management responses? Are these triggers different based on whether they result from a specific event (e.g., flood) or are ambient?
2. Did seagrass cover improve after the flood event(s)? If so, how and where?
3. Can we identify site-specific resilience indicators, which describe where recovery is likely to be enhanced or impeded?
4. What factors contribute to seagrass resilience?

The spatially explicit statistical models (i.e., spatial models) we used to answer these research questions are related to commonly used linear and generalized linear models but incorporate _spatial dependence_ Spatial dependence measures how similar two sites are based on their distance. Formally, we can build spatial dependence directly into spatial models by leveraging Tobler's First Law of Geography [@tobler1970computer], which states succinctly that "nearby sites tend to be more similar than distant sites". Not only is this intuitive based on "first principles" perspective, but spatial models to provide many advantages for modeling spatial data compared to models that assume independence among observations (i.e., nonspatial models). First, nonspatial models tend to underestimate uncertainty, yielding covariate standard errors and p-values for covariates that are too small and corresponding confidence intervals that are too narrow. Second, spatial models improve prediction at unsampled locations by borrowing strength from nearby observations, yielding more precise predictions and corresponding standard errors that vary by location. Third, spatial models decompose model error into separate spatial and nonspatial components, informing residual and resilience structure in ways that nonspatial models cannot. Fourth, spatial models can provide ecologically meaningful insight into the spatial dependence structure of the process, informing relationships among measured and unmeasured variables as well as future sampling plans. @zimmerman2024spatial provide a thorough review of spatial models for ecological and environmental data, providing more detail and justification for the claims made above.

Spatial models offer many benefits over commonly used machine learning algorithms like random forests [@breiman2001random; @cutler2007random; @fox2017assessing; @james2017introduction; @kuhn2022tidy]. First, traditional machine learning algorithms do not incorporate spatial dependence and, by consequence, often perform worse than spatial models at characterizing important covariates and predicting at unobserved locations [@fox2020comparing]. Second, machine learning algorithms struggle to quantify uncertainty in a principled manner, something spatial models excel at. Third, covariate inference is more straightforward using spatial models, as they naturally provide p-values that enable structured testing of hypotheses. Fourth, spatial models provide a rich set of diagnostics like leverage and Cook's distances [@montgomery2021introduction], quantities not generally defined for machine learning algorithms. This is not to say that machine learning algorithms cannot be very useful in ecological data analyses, but rather than spatial models are often a more effective tool for modeling spatial data. An exciting and developing area of statistical research involves incorporating spatial modeling strategies directly into machine learning algorithms [@saha2023random; @heaton2025scalable].

The first spatially explicit model we define is the spatial linear model. The spatial linear model is a _mixed-effects model_ [@pinheiro2000mixed; @bolker2015linear] written as

$$
y_i = \beta_0 + \beta_1x_{1, i} + \beta_2x_{2, i} + ... + \beta_p x_{p, i} + \tau_i + \epsilon_i,
$$ {#eq-splm1}
where $i = 1, 2, ..., n$ indexes the $n$ observations. The quantity $y_i$ is the response variable (for observation $i$), the $\beta$ parameters are slope parameters controlling the covariates, $x_{j, i}$ (for $j = 1, 2, ..., p$). The random error $\tau_i$ is a spatial random error, and the random error $\epsilon_i$ is the nonspatial (i.e., independent) random error. The spatial random error is spatially structured, capturing spatial residual effects like environmental gradients we do not have data to represent. The nonspatial random error is not spatially structure and captures nonspatial residual effects like measurement error and microscale variation. The difference between a traditional (i.e., nonspatial) linear model and a spatial linear model is the inclusion of $\tau_i$ (in the spatial linear model). The spatial random error implied by $\tau_i$ is governed by a _spatial covariance function_. The spatial covariance function allows the model to incorporate spatial dependence, yielding the improved model performance previously described. @fig-spcov_fns shows three different spatial covariance functions that describe different ways the spatial covariance decays with distance. Model parameters and standard errors are estimated from data using likelihood-based [@patterson1971recovery; @harville1977maximum; @wolfinger1994computing] or semivariogram-based [@cressie1985fitting; @curriero1999composite] estimation methods. Often, the spatial linear model in @eq-splm1 in is written more compactly using _matrix notation_:
$$
\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
$$ {#eq-splm2}
where $\mathbf{X}$ is a matrix that holds the $j$ covariates for each of $n$ observations, and $\mathbf{y}$, $\boldsymbol{\beta}$, $\boldsymbol{\tau}$, and $\boldsymbol{\tau}$ are vectors that contain each of the $n$ or $j$ quantities from @eq-splm1. For more information about specific forms of various spatial covariance functions, see @schabenberger2017statistical and @zimmerman2024spatial.

```{r}
#| label: fig-spcov_fns
#| fig-cap: "Three spatial covariance functions that represent different spatial dependence behavior."
range <- 50
spcor_dat <- tibble(
  h = seq(0, range*2),
  Exponential = exp(-h/range),
  Gaussian = exp(-h^2/range^2),
  Spherical = (1 - 1.5*h/range + 0.5*h^3/range^3)*(h <= range)
) %>%
  rename(Distance = h) %>%
  pivot_longer(cols = c(-Distance), names_to = "Function", values_to = "Spatial Covariance")
ggplot(spcor_dat, aes(x = Distance, y = `Spatial Covariance`, color = Function, linetype = Function)) +
  geom_line(linewidth = 2) +
  theme_bw(base_size = 14)
```

Like the traditional linear model, the spatial linear model can capture complex nonlinearities in the data through the use of interaction effects [@faraway2002practical; @lenth2025emmeans], splines [@chambers1992statistical], additive models [@wood2017generalized; @pedersen2019hierarchical], and penalized regression models like the ridge [@hastie2009elements] and lasso [@tibshirani1996regression]. Generally, this is accomplished by leveraging _basis functions_ that break down nonlinear behavior into a set of additive linear components that may include penalty terms which guard against overfitting. Furthermore, there is no assumption that $y_i$ itself is normally distributed; rather, distributional assumptions are placed on the random errors after incorporating the slope parameters. This clarification implies the spatial (and nonspatial) linear models can be quite effective at modeling a range of data that are not normally distributed. For highly nonnormal data, tools like generalized linear models, which we discuss later, are helpful. Finally, under general conditions, the slope parameters of the spatial (and nonspatial) linear models are approximately normally distributed with a large sample size (see @casella2024statistical for a general justification via the Central Limit Theorem and @zhang2005towards for nuances regarding spatial data). This is beneficial because it justifies valid hypothesis testing on the slope coefficients (often of primary ecological interest), no matter the distribution of the data (given certain assumptions are satisfied).

One of the most useful applications of the spatial linear model is predicting the response variable at unobserved locations. A very useful spatial prediction method called Kriging has a rich theoretical justification [@cressie1990origins; @stein1999interpolation]. Kriging is the spatial analogue to best linear unbiased prediction (BLUP) for mixed models [@henderson1975best]. Via Kriging, the spatial linear model produces BLUPs at each unobserved location of interest, alongside prediction intervals to characterize uncertainty. While useful, sometimes the goal is to create a BLUP for an entire region of interest. This technique, which builds upon the theoretical foundations of Kriging, is called block Kriging [@ver2002sampling; @cressie2006block]. The intution is to make point predictions at a very dense grid in the study region and then average these predictions, appropriately characterizing uncertainty. Together, Kriging and block Kriging provide a complete set of tools for spatial prediction, given the fit of a spatial linear model.

Spatial GLMs

spmodel, large data, glms, etc.

# Moreton Bay

## The Data

```{r}
#| label: read-mb
dat <- read_sf(here("data", "MBsub_03-31-2026.gpkg"))

# read in bay shapefiles
bay <- read_sf(here("data", "Moreton_Bay_marine_park_zoning_2008.shp")) %>%
  st_transform(crs = 3113)
```

We started with 11,796 seagrass cover observations in Moreton Bay across three financial years (2020-2021, 2022-2023, and 2023-2024). eReefs data [@steven2019ereefs] with modeled water quality covariates related to turbidity and salinity were only available at 8,364 observaations. Because one of our research questions involved connecting water quality covariates to management actions, we proceed with this subset of 8,364 observations for modeling. Of the 8,364 observations, 1,995 were in 2020-2021 (pre-flood), 4,036 were in 2022-2023, and 2,333 were in 2023-2024 (@tbl-n_fy_mb). Total cover is a percentage that ranges from 0 (no seagrass) to 100 (seagrass entirely contained in an area). @fig-cover_fy_mb shows total cover by financial year throughout the Bay. Total cover tends to be largest along the eastern shoreline of the Bay and lowest in the middle of the Bay. @fig-cover_fy_mb also reveals that 2022-2023 had by far the most sampling in the middle of the Bay.

```{r}
#| label: tbl-n_fy_mb
#| fig-cap: "Total seagrass cover in Moreton Bay by financial year."
dat %>%
  tabyl(fy) %>%
  adorn_totals(where = "row") %>%
  adorn_pct_formatting() %>%
  kable(
    col.names = c("Financial Year", "Sample Size", "Percent (of Total)"),
    format.args = list(big.mark = ",")
  )
```


```{r}
#| label: fig-cover_fy_mb
#| fig-cap: "Total seagrass cover in Moreton Bay by financial year."
dat <- dat %>%
  arrange(total.cover, fy)
ggplot(dat) +
  geom_sf(data = bay, fill = "white") +
  geom_sf(data = dat, aes(color = total.cover), size = 0.8) +
  scale_color_viridis_c(name = "Cover") +
  facet_wrap(~ fy) +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

We considered for modeling several water quality and geographic covariates. @tbl-wq_vars_mb summarize these water quality variables, which included both ambient and flood-related metrics related to turbidity, salinity, and temperature. The "Variable" column indicates the water quality variable turbidity, salinity, or temperature. The "Flood" variable indicates whether the measurement corresponds to a flood variable or not. The "Period" variable indicates the prior period (in days) relative to sampling or indicates whether the variable corresponds to a flood event. The "Type" column is the type of statistic, either a mean, minimum, maximum, number of days above (or below) a cutoff during "Period", or a maximum spell (i.e., duration) above (or below) a cutoff during "Period". The "Included" variable indicates whether the variable was included as a covariate in the final spatial model.

```{r}
#| label: tbl-wq_vars_mb
#| fig-cap: "Moreton Bay water quality variables considered for modeling."
turb_vars <- tibble(
  Variable = rep("Turbidity", 20),
  Flood = c(rep("No", 12), rep("Yes", 8)),
  Period = c(rep(c("7", "30", "90"), times = 4), rep("Flood", 8)),
  Type = c(rep("Mean", 3), rep("Above", 9), rep("Above", 3), rep("Duration", 3), c("Mean", "Max")),
  Cutoff = c(rep("NA", 3), rep(c("3", "5", "6"), each = 3), rep(c("3", "5", "6"), times = 2), rep("NA", 2))
) %>%
  mutate(Included = if_else(row_number() %in% c(3, 19), "Yes", "No"))
sal_vars <- tibble(
  Variable = rep("Salinity", 17),
  Flood = c(rep("No", 9), rep("Yes", 8)),
  Period = c(rep(c("7", "30", "90"), times = 3), rep("Flood", 8)),
  Type = c(rep("Below", 9), rep("Below", 3), rep("Duration", 3), c("Mean", "Min")),
  Cutoff = c(rep(c("20", "25", "30"), each = 3), rep(c("20", "25", "30"), times = 2), rep("NA", 2))
) %>%
  mutate(Included = if_else(row_number() %in% c(16), "Yes", "No"))
temp_vars <- tibble(
  Variable = rep("Temperature", 3),
  Flood = c(rep("No", 3)),
  Period = c(rep(c("7", "30", "90"), times = 1)),
  Type = c(rep("Mean", 3)),
  Cutoff = c(rep("NA", 3))
) %>%
  mutate(Included = if_else(row_number() %in% c(3), "Yes", "No"))
vars <- bind_rows(turb_vars, sal_vars, temp_vars)
vars %>%
  kable()
```


```{r}
set.seed(0)
form <- total.cover ~ sed.type + depth.range + temp.90d.mn + turb.90d.mn +
    after.flood + yr.flood + turb.fld.mn:yr.flood.recip
spmod <- splm(form, data = dat, spcov_type = "exponential", local = list(size = 500), random = ~ ehmp.subzone)
```

```{r}
formrf <- total.cover ~ sed.type + depth.range + temp.90d.mn + turb.90d.mn +
    after.flood + yr.flood + turb.fld.mn + ehmp.subzone
rf <- ranger(formrf, data = st_drop_geometry(dat), importance = "impurity")
```


# Wide Bay


## Outstanding Questions

* What are the units for the turbidity/salinity/temperature variables?

## References

::: {#refs}
:::